### 2016-09-18 ms.tex
start paper

\documentclass[manuscript, letterpaper]{aastex6}
\bibliographystyle{aasjournal}

\usepackage{graphicx}
\usepackage[suffix=]{epstopdf}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{url}
\usepackage{xspace}

% from here: https://github.com/dfm/peerless/blob/master/document/ms.tex#L19-L69
% ----------------------------------- %
% start of AASTeX mods by DWH and DFM %
% ----------------------------------- %

\setlength{\voffset}{0in}
\setlength{\hoffset}{0in}
\setlength{\textwidth}{6in}
\setlength{\textheight}{9in}
\setlength{\headheight}{0ex}
\setlength{\headsep}{\baselinestretch\baselineskip} % this is 2 lines in ``manuscript''
\setlength{\footnotesep}{0in}
\setlength{\topmargin}{-\headsep}
\setlength{\oddsidemargin}{0.25in}
\setlength{\evensidemargin}{0.25in}

\linespread{0.54} % close to 10/13 spacing in ``manuscript''
\setlength{\parindent}{0.54\baselineskip}
\hypersetup{colorlinks = false}
\makeatletter % you know you are living your life wrong when you need to do this
\long\def\frontmatter@title@above{
\vspace*{-\headsep}\vspace*{\headheight}
\noindent\footnotesize
{\noindent\footnotesize\textsc{\@journalinfo}}\par
{\noindent\scriptsize Preprint typeset using \LaTeX\ style AASTeX6\\
With modifications by David W. Hogg and Daniel Foreman-Mackey
}\par\vspace*{-\baselineskip}\vspace*{0.625in}
}%
\makeatother

% Section spacing:
\makeatletter
\let\origsection\section
\renewcommand\section{\@ifstar{\starsection}{\nostarsection}}
\newcommand\nostarsection[1]{\sectionprelude\origsection{#1}}
\newcommand\starsection[1]{\sectionprelude\origsection*{#1}}
\newcommand\sectionprelude{\vspace{1em}}
\let\origsubsection\subsection
\renewcommand\subsection{\@ifstar{\starsubsection}{\nostarsubsection}}
\newcommand\nostarsubsection[1]{\subsectionprelude\origsubsection{#1}}
\newcommand\starsubsection[1]{\subsectionprelude\origsubsection*{#1}}
\newcommand\subsectionprelude{\vspace{1em}}
\makeatother

\widowpenalty=10000
\clubpenalty=10000

\sloppy\sloppypar

% ------------------ %
% end of AASTeX mods %
% ------------------ %

%    Make Scientific Notation
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}}

% make the word Kepler italicized
\newcommand{\Kepler}{\textsl{Kepler}\xspace}

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%
\title{Demonstrating the Stellar Rotation \Kepler--Gaia }

\shorttitle{Rotating Stars}
\shortauthors{Davenport}

\author{
	James R. A. Davenport\altaffilmark{1,2}
	}

%\altaffiltext{1}{Corresponding author: James.Davenport@wwu.edu}
\altaffiltext{1}{Department of Physics \& Astronomy, Western Washington University, 516 High St., Bellingham, WA 98225, USA}
\altaffiltext{2}{NSF Astronomy and Astrophysics Postdoctoral Fellow}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Rotating stars. 
overlap between samples
two 
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

two period distribution found for M dwarfs in \citet{mcquillan2013}, confirmed in up to K dwarfs \citet{mcquillan2014} for full

%%%%%%%%%%%%%%%%%%%%%%
\section{data}

\Kepler mission \citep{borucki2010}

the Gaia mission \citep{gaia}, Data Release 1 \citep{gaia_dr1}

using CDS X-Match service
tgas = J/ApJS/211/24/table1 \& I/337/tgas
gaia = J/ApJS/211/24/table1 \& I/337/gaia

%%%%%%%%%%%%
\section{Determining Flare Energies}

\begin{figure}[]
\centering
\includegraphics[width=6in]{fig1.png}
\caption{
}
\label{fig:all}
\end{figure}

%%%%%%%%%%%%
\section{Determining Flare Energies}

\begin{figure}[]
\centering
\includegraphics[width=5.3in]{fig2.png}
\caption{
isochrones from \citet{bressan2012}
}
\label{fig:all}
\end{figure}

%%%%%%%%%%%%
\section{Determining Flare Energies}

\begin{figure}[]
\centering
\includegraphics[width=4in]{fig3.png}
\caption{
}
\label{fig:all}
\end{figure}

%%%%%%%%%%%%
\section{Determining Flare Energies}

\begin{figure}[]
\centering
\includegraphics[width=5.3in]{fig4.png}
\caption{
}
\label{fig:all}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
gaia useful for dtrrmining the age distribution of the nearby field
have shown utility for using Gaia data, combined with detailed light curve statitiscs from kepler, to reveal hidden structure in properties of field stars

the bimodality may be another manifestation of the ``Vaughan Preston gap'' \citep{vaughan1980}, e.g. discussed for rotating stars by\citet{kado-fong2016}.
either due to fast evolution thorugh intermediate stellar activity, or could be an age gap.

another way the full Gaia release could further contribute to this mystery is to model the star formation history of the stars in the \Kepler field, as well as the whole Milky Way \citep[e.g.][]{bertelli1999}. this will help A) improve gyrochronology relations by selecting only main sequence stars, and B) independently constrain age distribution for rotating stars to see if this gap is confirmed.

%%%%%%%%%%%%%%%%%
\acknowledgments
JRAD is supported by an NSF Astronomy and Astrophysics Postdoctoral Fellowship under award AST-1501418.

This research made use of the cross-match service provided by CDS, Strasbourg.

\bibliography{/Users/james/Dropbox/references}

\end{document}

### 2016-09-21 paper/ms.tex
add paper

--- 2016-09-18 ms.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%
-\title{Demonstrating the Stellar Rotation \Kepler--Gaia }
+\title{Rotating Stars from \Kepler Observed with Gaia}
 
@@
 overlap between samples
-two 
+
 \end{abstract}
@@
 
+of course, Kepler useful for calibrations for Gaia, e.g. log(g) \citep{creevey2013}
+
 two period distribution found for M dwarfs in \citet{mcquillan2013}, confirmed in up to K dwarfs \citet{mcquillan2014} for full
-
-%%%%%%%%%%%%%%%%%%%%%%
-\section{data}
 
@@
 
-using CDS X-Match service
-tgas = J/ApJS/211/24/table1 \& I/337/tgas
-gaia = J/ApJS/211/24/table1 \& I/337/gaia
+%%%%%%%%%%%%%%%%%%%%%%
+\section{The \Kepler--Gaia Data}
 
-%%%%%%%%%%%%
-\section{Determining Flare Energies}
+using CDS X-Match service, we match the catalogs from 
+%tgas = J/ApJS/211/24/table1 \& I/337/tgas
+%gaia = J/ApJS/211/24/table1 \& I/337/gaia
 
@@
 %%%%%%%%%%%%
-\section{Determining Flare Energies}
+\section{Selecting Main Sequence Stars}
+
+first we create the HR diagram, using the absolute Gaia magnitudes from the TGAS parallaxes, and the temperatures reported from McQ14
 
@@
 }
-\label{fig:all}
+\label{fig:HR}
 \end{figure}
-
-%%%%%%%%%%%%
-\section{Determining Flare Energies}
 
@@
 }
-\label{fig:all}
+\label{fig:hist}
 \end{figure}
@@
 %%%%%%%%%%%%
-\section{Determining Flare Energies}
+\section{Extending the Spin-Down Gap}
+
+this feature first observed in rotation period by \citet{mcquillan2013} for M dwarfs
+
+here we show it appears to extended to nearby higher mass stars in the \Kepler field.
+
+the reason this feature did not appear in earlier \Kepler studies is the inability to select main sequence stars from turn-off or subgiants using the KIC
 
@@
 }
-\label{fig:all}
+\label{fig:gyro}
 \end{figure}

### 2016-10-14 paper/ms.tex
split isochrone
real paper outlining happening

@@
 \setlength{\parindent}{0.54\baselineskip}
-\hypersetup{colorlinks = false}
+%\hypersetup{colorlinks = false}
 \makeatletter % you know you are living your life wrong when you need to do this
@@
 %%%%%%%%%%%%%%%%%%%%%%
-\title{Rotating Stars from \Kepler Observed with Gaia}
+\title{Rotating Stars from \Kepler Observed with Gaia DR1}
 
@@
 \begin{abstract}
-Rotating stars. 
-overlap between samples
+Astrometric data from the recent Gaia Data Release 1 has been matched against the sample of stars from \Kepler with known rotation periods.
+A total of 1,303 bright stars were recovered from the subset of Gaia sources with good astrometric solutions, most with temperatures hotter than 5000 K.
 
@@
 
-of course, Kepler useful for calibrations for Gaia, e.g. log(g) \citep{creevey2013}
+\Kepler mission \citep{borucki2010} has enabled statistical study of rotation periods for field stars for the first time.
+This fundamental stellar property has been measured for over 30k stars by tracing the periodic or quasi-periodic modulations of their light curves as cool starspots rotate in and out of view \citep{reinhold2013,mcquillan2014}.
+The seminal work by \citet{skumanich1972} connected stellar rotation and age via angular momentum loss, leading to an age estimating technique known as gyrochronology.
+With rotation now being routinely measured, it is hoped a new ability of determining robust ages for field stars will be possible by calibrating gyro-isochrones, and  \citep[e.g.][]{angus2015,van-saders2016}.
 
-two period distribution found for M dwarfs in \citet{mcquillan2013}, confirmed in up to K dwarfs \citet{mcquillan2014} for full
+Curiously, two period distributions found for M dwarfs in \citet{mcquillan2013}, confirmed in up to K dwarfs \citet{mcquillan2014}.
+Why this feature is not found at hotter temperatures, or has not been seen in other comprehensive rotation period studies (such as in open clusters), is a mystery.
+Currently favored explanations are an age distribution (CITE) or a rapid evolution through this period, similar to the Vaughan-Preston gap (?).
+Binary stars and period contamination by evolved stars may also be confounding our understanding of the period distribution seen for \Kepler targets.
 
-\Kepler mission \citep{borucki2010}
+astrometric data from the Gaia mission \citep{gaia} can help shed light on this mystery.
+By measuring distances via stellar parallax, the \Kepler--Gaia sample can separate single main sequence dwarfs from binary stars or evolved stars such as sub-giants, and can help calibrate fundamental properties of \Kepler stars, such as log(g) \citep{creevey2013}.
+This will help isolate the cause of the period bimodality,
 
-the Gaia mission \citep{gaia}, Data Release 1 \citep{gaia_dr1}
+In this paper I demonstrate the utility of combining \Kepler rotation period sample with the preliminary Data Release 1 from Gaia \citep{gaia_dr1}.
+This combined sample allows improved selection of main sequence stars, and reveals previously undetected structure in the rotation period distribution for solar-type stars.
 
@@
 \section{The \Kepler--Gaia Data}
+Rotation periods in this study come from \citet{mcquillan2014}, who performed an Auto-Correlation Function analysis of \Kepler stars cooler than 6500 K that had at least $\sim$2 years of observation.
+The periods recovered from this approach generally agree very well with those found via Lomb-Scargle Periodograms \citep[e.g.][]{aigrain2015}.
+Sources with multiple distinct periods, such as from binary systems with two spotted stars \citep[e.g.][]{lurie2015} are detected by \citet{mcquillan2014}, but are not included in the following analysis.
 
-using CDS X-Match service, we match the catalogs from 
+The Gaia Data Release 1 (DR1) provides astrometric positions for over $10^9$ sources from the first year of observation with Gaia.
+The Tycho-Gaia Astrometric Solution (TGAS) measures improved proper motions and parallaxes for 2 million nearby, bright sources by extending the astrometric solutions from Tycho and Hipparcos.
+
+using CDS X-Match service, I matched the catalogs from these two surveys.
+The rotation period distribution is shown in Figure \ref{fig:all}.
+A total of 41,739 \Kepler rotation period sources were found in the Gaia DR1 catalog.
+A subset of 1,303 objects were recovered in the TGAS sample.
+Due to the brightness limits of the TGAS sample very few K and M dwarfs were recovered in the TGAS sample.
+Future releases of Gaia data will ostensibly provide astrometry for nearly all \Kepler stars.
+
 %tgas = J/ApJS/211/24/table1 \& I/337/tgas
@@
 \caption{
+Rotation period distribution for 41,739 \Kepler stars from \citet{mcquillan2014} with detections in Gaia DR1 (blue dots).
+The period bimodality can be seen most clearly for stars with $T_{eff} < 4000$ K as a dearth of sources with periods of $\sim$20 days, but extends to at least $T_{eff}\sim5500$ K according to \citet{mcquillan2014}.
+The subsample of 1,303 nearby objects found in TGAS are highlighted (red circles), and are mostly hotter stars due to the faint limit of the TGAS sample.
 }
@@
 \includegraphics[width=5.3in]{fig4.png}
-\caption{
+\caption{ Rotation period versus temperature for TGAS-matched stars near the isochrone main sequence (blue circles).
+Distributions of rotation periods for these main sequence selected stars are shown as vertical histograms for 7 bins of temperature (black lines), with each temperature bin having equal numbers of stars.
+As in Figure \ref{fig:all}, the full \Kepler--Gaia matched sample is shown for reference (small black dots).
+The bimodality in rotation periods previously seen by \citet{mcquillan2014} extends the full range of temperatures in the \Kepler--Gaia main sequence sample shown here.
+A \citet{meibom2011} 600 Myr gyrochonology-isochrone roughly traces the bimodality midpoint up to 6000 K (red solid line).
+A log-linear extrapolation to 6500 K (red dashed line) continues to track the bimodality to hotter temperatures, and roughly traces a line of constant Rossby number.
 }
@@
 \section{Discussion}
-gaia useful for dtrrmining the age distribution of the nearby field
-have shown utility for using Gaia data, combined with detailed light curve statitiscs from kepler, to reveal hidden structure in properties of field stars
+gaia useful for determining the age distribution of the nearby field
+have shown utility for using Gaia data, combined with detailed light curve statitiscs from kepler, to reveal hidden structure in properties of field stars.
+This combination will be super useful for determining approximate log g values (separating main sequence from sub-giants), and thus more accurate gyrochronology ages \citep{van-saders2013}
+
+another way the full Gaia release could further contribute to this mystery is to model the star formation history of the stars in the \Kepler field, as well as the whole Milky Way \citep[e.g.][]{bertelli1999}.
+This was previously tried using the color--absolute magnitude diagram from Hipparcos by \citet{hernandez2000}, who reconstructed a non-continuous star formation history for the local solar neighborhood.
 
 the bimodality may be another manifestation of the ``Vaughan Preston gap'' \citep{vaughan1980}, e.g. discussed for rotating stars by\citet{kado-fong2016}.
-either due to fast evolution thorugh intermediate stellar activity, or could be an age gap.
+either due to fast evolution through intermediate stellar activity, or could be an age gap.
+if age, these clumps line up roughly with a 300 Myr and 2 Gyr \citet{meibom2011} gyro-isochrone from cooler than about 6000 K
 
-another way the full Gaia release could further contribute to this mystery is to model the star formation history of the stars in the \Kepler field, as well as the whole Milky Way \citep[e.g.][]{bertelli1999}.
-this will help A) improve gyrochronology relations by selecting only main sequence stars, and B) independently constrain age distribution for rotating stars to see if this gap is confirmed.
+median distance of K/M stars ($T_{eff} > 4000$ K) is $\sim$216 pc, using isochrone
+median distance for the TGAS-matched sample (blue points) is very close, 285 pc, so also sampling just the most local volume.
+this points to the effect being localized around us, which explains why it has not been seen in other gyro studies to date (e.g. clusters)
+Unfortunately not enough rotation period data across the fully convective boundary ($\sim$3000 K) to tell if this bimodal feature continues.
 
@@
 \acknowledgments
+Special thanks to Jennifer van Saders, Sean Matt, and Travis Metcalfe for their helpful discussions that motivated publishing this work.
+
 JRAD is supported by an NSF Astronomy and Astrophysics Postdoctoral Fellowship under award AST-1501418.

### 2016-10-14 paper/ms.tex
adding more text

@@
 
+% (gaia[u'parallax_error'] < 0.4) &
+  %            (gaia[u'phot_g_mean_flux_error']/gaia[u'phot_g_mean_flux'] < 0.01))
+
 first we create the HR diagram, using the absolute Gaia magnitudes from the TGAS parallaxes, and the temperatures reported from McQ14
 
+%%%%%
 \begin{figure}[]
@@
 \includegraphics[width=5.3in]{fig2.png}
-\caption{
-isochrones from \citet{bressan2012}
+\caption{Hertzsprung--Russell (HR) diagram using temperatures from \citet{mcquillan2014} and Gaia DR1 $G$-band absolute magnitudes for the 894 stars that pass photometric and parallax quality cuts described in the text.
+Points are colored by their measured \Kepler rotation periods.
+Two isochrones from \citet{bressan2012} are shown, with ages of 300 Myr and 4 Gyr (solid and dashed lines).
 }
@@
 
+%%%%%
 \begin{figure}[]
@@
 \includegraphics[width=4in]{fig3.png}
-\caption{
-}
+\caption{Distribution of differences between the Gaia DR1 $G$ absolute magnitudes and values for main sequence stars from the 300 Myr isochrone shown in Figure \ref{fig:HR}.
+The range of sources selected near the single star main sequence (red lines) may have additional contamination due to dust extinction and binary companions.}
 \label{fig:hist}

### 2016-10-16 paper/ms.tex
a couple new plots

@@
 \centering
-\includegraphics[width=5.3in]{fig4.png}
+\includegraphics[width=5.3in]{fig4alt.png}
 \caption{ Rotation period versus temperature for TGAS-matched stars near the isochrone main sequence (blue circles). 
-Distributions of rotation periods for these main sequence selected stars are shown as vertical histograms for 7 bins of temperature (black lines), with each temperature bin having equal numbers of stars.
+%Distributions of rotation periods for these main sequence selected stars are shown as vertical histograms for 7 bins of temperature (black lines), with each temperature bin having equal numbers of stars.
 As in Figure \ref{fig:all}, the full \Kepler--Gaia matched sample is shown for reference (small black dots).
 The bimodality in rotation periods previously seen by \citet{mcquillan2014} extends the full range of temperatures in the \Kepler--Gaia main sequence sample shown here.
-A \citet{meibom2011} 600 Myr gyrochonology-isochrone roughly traces the bimodality midpoint up to 6000 K (red solid line).
-A log-linear extrapolation to 6500 K (red dashed line) continues to track the bimodality to hotter temperatures, and roughly traces a line of constant Rossby number.
+A \citet{meibom2011} 600 Myr gyrochonology-isochrone roughly traces the bimodality midpoint up to 6000 K (red solid line), but deviates sharply up to $\sim$6200 K (red dotted line).
+A log-linear extrapolation to of the isochrone from 6000K to 6500 K (red dashed line) continues to track the bimodality to hotter temperatures, and roughly traces a line of constant Rossby number.
 }
@@
 
+\begin{figure}[]
+\centering
+\includegraphics[width=4in]{fig5.png}
+\caption{Residual of log rotation periods about the \citet{meibom2011} 600 Myr isochrone, using the log-linear extrapolation between 6000--6500 K shown in Figure \ref{fig:gyro}.
+}
+\label{fig:diff}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%
 \section{Discussion}
-gaia useful for determining the age distribution of the nearby field
-have shown utility for using Gaia data, combined with detailed light curve statitiscs from kepler, to reveal hidden structure in properties of field stars.
-This combination will be super useful for determining approximate log g values (separating main sequence from sub-giants), and thus more accurate gyrochronology ages \citep{van-saders2013}
-
-another way the full Gaia release could further contribute to this mystery is to model the star formation history of the stars in the \Kepler field, as well as the whole Milky Way \citep[e.g.][]{bertelli1999}.
-This was previously tried using the color--absolute magnitude diagram from Hipparcos by \citet{hernandez2000}, who reconstructed a non-continuous star formation history for the local solar neighborhood.
-
-the bimodality may be another manifestation of the ``Vaughan Preston gap'' \citep{vaughan1980}, e.g. discussed for rotating stars by\citet{kado-fong2016}.
-either due to fast evolution through intermediate stellar activity, or could be an age gap.
-if age, these clumps line up roughly with a 300 Myr and 2 Gyr \citet{meibom2011} gyro-isochrone from cooler than about 6000 K
 
@@
 Unfortunately not enough rotation period data across the fully convective boundary ($\sim$3000 K) to tell if this bimodal feature continues.
+
+the bimodality may be another manifestation of the ``Vaughan Preston gap'' \citep{vaughan1980}, e.g. discussed for rotating stars by\citet{kado-fong2016}.
+either due to fast evolution through intermediate stellar activity, or could be an age gap.
+if age, these clumps line up roughly with a 300 Myr and 2 Gyr \citet{meibom2011} gyro-isochrone from cooler than about 6000 K.
+this similar to the non-continusous star formation history for the local solar neighborhood using Hipparcos astrometry \citep{hernandez2000}.
+this age explanation was bolstered in \citet{mcquillan2013}, who noted the two populations of rotating M dwarfs had different distributions of proper motions, indicating they belonged to distinct groups of stars.
+The distribution of total proper motion for stars above and below the modified 600 Myr gyrochrone is shown in Figure \ref{fig:pm}.
+Stars above the gyrochrone (slower rotators, nominally older) have a median total proper motion of 15.4 mas/yr, while those below (faster rotators, younger)  have a median of 11.3 mas/yr.
+This difference in kinematics versus rotation period is in the same direction observed by \citet{mcquillan2013}.
+As the typical error in the total proper motion is $\sim$1.4 mas/yr for this sample, this is a marginal (2.8$\sigma$) difference.
+Note also that likely contamination of subgiants in the hotter stars may also obscure kinematic differences.
+
+%('median above: ', 15.373856668858519)
+%('median below: ', 11.301950163653459)
+%('median PM error:', 1.4275539188520987)
+%('diff / error = ', 2.8523661708548951)
+\begin{figure}[]
+\centering
+\includegraphics[width=4in]{fig6.png}
+\caption{Total proper motion distributions for stars above (rotating slower, older stars) the gyrochrone model shown in Figure \ref{fig:gyro} (black) and below (rotating faster, younger stars) the model (green).
+Median values of the two distributions are shown (thick lines), which yield a marginal 2.8$\sigma$ difference.
+}
+\label{fig:pm}
+\end{figure}
+
+finally, %gaia + \Kepler useful for determining the age distribution of the nearby field
+have shown utility for using Gaia data, combined with detailed light curve statitiscs from \Kepler, to reveal hidden structure in properties of field stars.
+This combination will be super useful for determining approximate log g values (separating main sequence from sub-giants), and thus more accurate gyrochronology ages \citep{van-saders2013}.
+also for modeling the star formation history of the entire milky way \citep[e.g.][]{bertelli1999}.
 

### 2016-10-18 paper/ms.tex
writing from GaiaSprint day 2

@@
 \documentclass[manuscript, letterpaper]{aastex6}
+%\documentclass[twocolumn]{aastex6}
+
 \bibliographystyle{aasjournal}
@@
 % ----------------------------------- %
-
 \setlength{\voffset}{0in}
@@
 \setlength{\evensidemargin}{0.25in}
-
 \linespread{0.54} % close to 10/13 spacing in ``manuscript''
@@
 \makeatother
-
 % Section spacing:
@@
 \makeatother
-
 \widowpenalty=10000
 \clubpenalty=10000
-
 \sloppy\sloppypar
-
 % ------------------ %
@@
 Astrometric data from the recent Gaia Data Release 1 has been matched against the sample of stars from \Kepler with known rotation periods.
-A total of 1,303 bright stars were recovered from the subset of Gaia sources with good astrometric solutions, most with temperatures hotter than 5000 K.
-
+A total of 1,299 bright rotating stars were recovered from the subset of Gaia sources with good astrometric solutions, most with temperatures hotter than 5000 K.
+From these sources, 894 were selected as lying near the main sequence using their absolute $G$-band magnitudes.
+These main sequence stars show a bimodality in their rotation period distribution, centered roughly around a 600 Myr rotation-isochrone.
+This feature matches the bimodal period distribution from cooler stars with \Kepler, but was previously undetected for solar-type stars due to sample contamination by subgiants.
+A tenuous connection between the rotation period and total proper motion is found, suggesting the period bimodality is due to the age distribution of stars within $\sim$300pc of the Sun, rather than a phase of rapid angular momentum loss.
+This emphasizes the unique power for stellar populations studies of combining temporal monitoring from \Kepler with astrometric data from Gaia
 \end{abstract}
@@
 
-\Kepler mission \citep{borucki2010} has enabled statistical study of rotation periods for field stars for the first time.
-This fundamental stellar property has been measured for over 30k stars by tracing the periodic or quasi-periodic modulations of their light curves as cool starspots rotate in and out of view \citep{reinhold2013,mcquillan2014}.
+The\Kepler mission \citep{borucki2010} has enabled the first studies of rotation periods for large ensembles of field stars.
+The fundamental stellar property of rotation has been measured for over 30k stars using the high cadence \Kepler light curves, tracing the periodic or quasi-periodic modulations in brightness as cool starspots rotate in and out of view \citep{reinhold2013,mcquillan2014}.
 The seminal work by \citet{skumanich1972} connected stellar rotation and age via angular momentum loss, leading to an age estimating technique known as gyrochronology.
+At present, ages determined by gyrochronology are accurate to $\sim$10\% in the {\it best} cases (young solar-type stars).
 With rotation now being routinely measured, it is hoped a new ability of determining robust ages for field stars will be possible by calibrating gyro-isochrones, and  \citep[e.g.][]{angus2015,van-saders2016}.
 
-Curiously, two period distributions found for M dwarfs in \citet{mcquillan2013}, confirmed in up to K dwarfs \citet{mcquillan2014}.
-Why this feature is not found at hotter temperatures, or has not been seen in other comprehensive rotation period studies (such as in open clusters), is a mystery.
-Currently favored explanations are an age distribution (CITE) or a rapid evolution through this period, similar to the Vaughan-Preston gap (?).
-Binary stars and period contamination by evolved stars may also be confounding our understanding of the period distribution seen for \Kepler targets.
+Curiously, in studying the rotation periods for M dwarfs in the \Kepler field, \citet{mcquillan2013} discovered a bimodal period distribution, which was subsequently confirmed to exist up to K dwarfs in the \Kepler field by \citet{mcquillan2014}.
+However, this had never been observed in any other study of stellar rotation periods, including stellar clusters at a variety of ages, nor was it detected in the \Kepler stars at hotter temperatures ($T_{eff} > 5000$).
+While binary stars and multiple-period systems may be contaminating the rotation period sample for \Kepler field stars, \citet{mcquillan2014} found they could not adequately explain the bimodal period distribution.
+Currently favored explanations for this feature are 1) a non-continuous age distribution for nearby stars, as was suggested with very nearby Hipparcos stars by \citet{hernandez2000}, or 2) a previously unknown phase of rapid angular momentum loss for low-mass stars, similar to the ``Vaughan-Preston'' gap seen in chromospheric activity indicators \citep{vaughan1980}.
+As independent age indicators for these field stars are often non-existent, and both scenarios deal with physical mechanisms that are not currently understood with precision, a definitive explanation has not been found.
 
-astrometric data from the Gaia mission \citep{gaia} can help shed light on this mystery.
-By measuring distances via stellar parallax, the \Kepler--Gaia sample can separate single main sequence dwarfs from binary stars or evolved stars such as sub-giants, and can help calibrate fundamental properties of \Kepler stars, such as log(g) \citep{creevey2013}.
-This will help isolate the cause of the period bimodality,
+Astrometric data from the Gaia mission \citep{gaia} can help shed light on this stellar population mystery.
+By measuring distances via stellar parallax for these rotating stars, the \Kepler--Gaia sample can separate single main sequence dwarfs from binary stars or evolved stars such as subgiants, and will help calibrate fundamental properties of \Kepler stars, such as log(g) \citep{creevey2013}.
+Galactic kinematics from Gaia will also provide an additional age-proxy, and allow for searches of substructure in field star ages such as from moving groups.
+The Gaia data will also enable a measurement of the star formation history of the disk from both white dwarf cooling sequences \citep{carrasco2014,gaensicke2015} and color-magnitude diagram models \citep{bertelli1999}.
 
-In this paper I demonstrate the utility of combining \Kepler rotation period sample with the preliminary Data Release 1 from Gaia \citep{gaia_dr1}.
+In this paper I demonstrate the utility of combining temporal properties derived from \Kepler light curves with the preliminary astrometric solutions from  Gaia Data Release 1 \citep[hereafter DR1][]{gaia_dr1}.
 This combined sample allows improved selection of main sequence stars, and reveals previously undetected structure in the rotation period distribution for solar-type stars.
@@
 Rotation periods in this study come from \citet{mcquillan2014}, who performed an Auto-Correlation Function analysis of \Kepler stars cooler than 6500 K that had at least $\sim$2 years of observation.
-The periods recovered from this approach generally agree very well with those found via Lomb-Scargle Periodograms \citep[e.g.][]{aigrain2015}.
+The periods recovered from this approach generally agree very well with those found via Lomb-Scargle Periodograms \citep[e.g.][]{reinhold2013,aigrain2015}.
 Sources with multiple distinct periods, such as from binary systems with two spotted stars \citep[e.g.][]{lurie2015} are detected by \citet{mcquillan2014}, but are not included in the following analysis.
@@
 The Tycho-Gaia Astrometric Solution (TGAS) measures improved proper motions and parallaxes for 2 million nearby, bright sources by extending the astrometric solutions from Tycho and Hipparcos.
+While the TGAS data are not a complete astrometric survey, and have possible systematics in the reported parallaxes \citep{stassun2016}, they represent a significant improvement in the astrometry and kinematics available for stars in the \Kepler field thanks to the precision of Gaia.
 
-using CDS X-Match service, I matched the catalogs from these two surveys.
-The rotation period distribution is shown in Figure \ref{fig:all}.
-A total of 41,739 \Kepler rotation period sources were found in the Gaia DR1 catalog.
-A subset of 1,303 objects were recovered in the TGAS sample.
+Using the CDS X-Match service, I cross-matched the available catalogs from these two surveys.
+A default cross-match radius of 5 arcsec was used.
+A total of 33,855 stars were found in cross match between these catalogs, 99.5\% of the sample from \citet{mcquillan2014}.
+A subset of 1,299 objects were recovered in the TGAS sample.
 Due to the brightness limits of the TGAS sample very few K and M dwarfs were recovered in the TGAS sample.
-Future releases of Gaia data will ostensibly provide astrometry for nearly all \Kepler stars.
+Future releases of Gaia data will ostensibly provide full astrometric solutions for nearly all \Kepler stars.
+The rotation periods versus stellar effective temperatures for the \Kepler--Gaia matched stars are shown in Figure \ref{fig:all}.
 
@@
 %gaia = J/ApJS/211/24/table1 \& I/337/gaia
-
 \begin{figure}[]
@@
 \caption{
-Rotation period distribution for 41,739 \Kepler stars from \citet{mcquillan2014} with detections in Gaia DR1 (blue dots).
+Rotation period distribution for 33,855 \Kepler stars from \citet{mcquillan2014} with detections in Gaia DR1 (blue dots).
 The period bimodality can be seen most clearly for stars with $T_{eff} < 4000$ K as a dearth of sources with periods of $\sim$20 days, but extends to at least $T_{eff}\sim5500$ K according to \citet{mcquillan2014}.
-The subsample of 1,303 nearby objects found in TGAS are highlighted (red circles), and are mostly hotter stars due to the faint limit of the TGAS sample.
+The subsample of 1,299 nearby objects found in TGAS are highlighted (red circles), and are mostly hotter stars due to the faint limit of the TGAS sample.
 }
@@
 
-%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%
 \section{Selecting Main Sequence Stars}
@@
   %            (gaia[u'phot_g_mean_flux_error']/gaia[u'phot_g_mean_flux'] < 0.01))
+Though \citet{mcquillan2014} attempted to only measure periods for dwarf starss, the sample of \Kepler--Gaia matched stars contains both main sequence dwarfs and evolved stars (giants and subgiants).
+Previous studies have shown sometimes significant contamination by giants can affect the implied variability properties for dwarf stars \citep{ciardi2011,mann2012}.
+Therefore to properly understand the nature of the period distribution, a sample of robust main sequence stars must be selected.
 
-first we create the HR diagram, using the absolute Gaia magnitudes from the TGAS parallaxes, and the temperatures reported from McQ14
+Faint stars were removed by requiring sources have $G$-band flux errors $<1$\%.
+To ensure accurate distances, and therefore luminosities, parallaxes were required to have errors $<0.4$ mas / year.
+These cuts left a total of 894 stars from the \Kepler--TGAS matched sample (68\%).
+The Hertzsprung--Russell (HR) diagram for these stars is shown in Figure \ref{fig:HR}.
 
@@
 \caption{Residual of log rotation periods about the \citet{meibom2011} 600 Myr isochrone, using the log-linear extrapolation between 6000--6500 K shown in Figure \ref{fig:gyro}.
+The bimodal rotation period distribution is clear, with peaks at $\Delta \log P_{rot}$ -0.19 and 0.21 days.
 }
@@
 
-\bibliography{/Users/james/Dropbox/references}
+\bibliography{/Users/davenpj3/Dropbox/references}
 

### 2016-10-19 paper/ms.tex
done for Tuesday

@@
 
-%%%%%%%%%%%%%%%%%%%%%%
-\section{Selecting Main Sequence Stars}
-
-% (gaia[u'parallax_error'] < 0.4) &
-  %            (gaia[u'phot_g_mean_flux_error']/gaia[u'phot_g_mean_flux'] < 0.01))
-Though \citet{mcquillan2014} attempted to only measure periods for dwarf starss, the sample of \Kepler--Gaia matched stars contains both main sequence dwarfs and evolved stars (giants and subgiants).
-Previous studies have shown sometimes significant contamination by giants can affect the implied variability properties for dwarf stars \citep{ciardi2011,mann2012}.
-Therefore to properly understand the nature of the period distribution, a sample of robust main sequence stars must be selected.
-
-Faint stars were removed by requiring sources have $G$-band flux errors $<1$\%.
-To ensure accurate distances, and therefore luminosities, parallaxes were required to have errors $<0.4$ mas / year.
-These cuts left a total of 894 stars from the \Kepler--TGAS matched sample (68\%).
-The Hertzsprung--Russell (HR) diagram for these stars is shown in Figure \ref{fig:HR}.
-
 %%%%%
@@
 \end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%
+\section{Selecting Main Sequence Stars}
+
+% (gaia[u'parallax_error'] < 0.4) &
+  %            (gaia[u'phot_g_mean_flux_error']/gaia[u'phot_g_mean_flux'] < 0.01))
+Though \citet{mcquillan2014} attempted to only measure periods for dwarf starss, the sample of \Kepler--Gaia matched stars contains both main sequence dwarfs and evolved stars (giants and subgiants).
+Previous studies have shown sometimes significant contamination by giants can affect the implied variability properties for dwarf stars \citep{ciardi2011,mann2012}.
+Therefore to properly understand the nature of the period distribution, a robust sample of main sequence stars must be selected.
+
+Faint stars were removed by requiring sources have $G$-band flux errors $<1$\%.
+To ensure accurate distances, and therefore luminosities, parallaxes were required to have errors $<0.4$ mas year$^{-1}$.
+These cuts left a total of 894 stars from the \Kepler--TGAS matched sample (68\%).
+The Hertzsprung--Russell (HR) diagram for these stars is shown in Figure \ref{fig:HR}, with each point colored by its \Kepler-measured rotation period.
+Example isochrones from the \citet{bressan2012} grid are shown for two ages.
+A systematic offset between the measured absolute $G$-band and the isochrone main sequence of $\sim$0.5 magnitudes is found (Figure \ref{fig:hist}).
+
+The HR-period diagram in Figure \ref{fig:HR} shows stars with a range of evolutionary states, and could help test post-main sequence angular momentum evolution models \citep[e.g.][]{donascimento2012}.
+Outliers in this diagram (e.g.
+the rapidly rotating star at $T_{eff}=4500$ K, $M_G=4.5$ mag, $P_{rot}$=0.652 d, KIC 07957709) are either due to erroneous cross-matching between the \Kepler and Gaia catalogs, or represent interesting systems such as rare binary star configurations or stars that have undergone mergers or ingested giant planets \citep{massarotti2008,tayar2015}.
+Explaining such outliers is beyond the scope of this work, but are worth further study.
+
+The difference between the 300 Myr isochrone main sequence track and the observed $M_G$ is shown in Figure \ref{fig:hist}.
+Given the systematic offset of $\sim$0.5 mag, a fairly wide band of stars ($0\le \Delta M_G \le 1$) was selected as being ``close to the main sequence''.
+This final sample included 440 stars.
+This simplistic cut is not a robust dwarf--giant, nor single--binary star separator, but serves to select a sample of mostly main sequence stars for the illustrative purpose of this work.
+More precise selection will require an improved isochrone track, as well as updated parallaxes from the full Gaia DR2.
 
@@
 \caption{Residual of log rotation periods about the \citet{meibom2011} 600 Myr isochrone, using the log-linear extrapolation between 6000--6500 K shown in Figure \ref{fig:gyro}.
-The bimodal rotation period distribution is clear, with peaks at $\Delta \log P_{rot}$ -0.19 and 0.21 days.
+The bimodal rotation period distribution is clear, with peaks at $\Delta \log P_{rot}$ of -0.19 and 0.21 days.
+A two-Gaussian model was fit to this distribution (blue line).
 }

### 2016-10-19 paper/ms.tex
Wednesday sprint writing done

@@
 
-The\Kepler mission \citep{borucki2010} has enabled the first studies of rotation periods for large ensembles of field stars.
+The \Kepler mission \citep{borucki2010} has enabled the first studies of rotation periods for large ensembles of field stars.
 The fundamental stellar property of rotation has been measured for over 30k stars using the high cadence \Kepler light curves, tracing the periodic or quasi-periodic modulations in brightness as cool starspots rotate in and out of view \citep{reinhold2013,mcquillan2014}.
@@
 At present, ages determined by gyrochronology are accurate to $\sim$10\% in the {\it best} cases (young solar-type stars).
-With rotation now being routinely measured, it is hoped a new ability of determining robust ages for field stars will be possible by calibrating gyro-isochrones, and  \citep[e.g.][]{angus2015,van-saders2016}.
+With rotation now being routinely measured, it is hoped a new ability of determining robust ages for field stars will be possible by calibrating gyro-isochrones to stellar clusters and asteroseismic samples, and improved models of angular momentum loss \citep[e.g.][]{angus2015,van-saders2016}.
 
@@
 Rotation period distribution for 33,855 \Kepler stars from \citet{mcquillan2014} with detections in Gaia DR1 (blue dots).
-The period bimodality can be seen most clearly for stars with $T_{eff} < 4000$ K as a dearth of sources with periods of $\sim$20 days, but extends to at least $T_{eff}\sim5500$ K according to \citet{mcquillan2014}.
+The period bimodality can be seen most clearly for stars with $T_{eff} < 4000$ K as a dearth of sources with periods of $\sim$25 days, but extends to at least $T_{eff}\sim5500$ K according to \citet{mcquillan2014}.
 The subsample of 1,299 nearby objects found in TGAS are highlighted (red circles), and are mostly hotter stars due to the faint limit of the TGAS sample.
@@
   %            (gaia[u'phot_g_mean_flux_error']/gaia[u'phot_g_mean_flux'] < 0.01))
-Though \citet{mcquillan2014} attempted to only measure periods for dwarf starss, the sample of \Kepler--Gaia matched stars contains both main sequence dwarfs and evolved stars (giants and subgiants).
-Previous studies have shown sometimes significant contamination by giants can affect the implied variability properties for dwarf stars \citep{ciardi2011,mann2012}.
-Therefore to properly understand the nature of the period distribution, a robust sample of main sequence stars must be selected.
+Though \citet{mcquillan2014} attempted to only measure periods for dwarf stars, the sample of \Kepler--Gaia matched stars contains both main sequence dwarfs and evolved stars (giants and subgiants).
+Previous studies have shown  significant contamination by giants or subgiants can affect the implied variability properties of dwarf stars \citep{ciardi2011,mann2012}.
+Therefore to properly understand the nature of the period distribution and its implications for age-dating field stars, a robust sample of main sequence stars must be selected.
 
@@
 the rapidly rotating star at $T_{eff}=4500$ K, $M_G=4.5$ mag, $P_{rot}$=0.652 d, KIC 07957709) are either due to erroneous cross-matching between the \Kepler and Gaia catalogs, or represent interesting systems such as rare binary star configurations or stars that have undergone mergers or ingested giant planets \citep{massarotti2008,tayar2015}.
-Explaining such outliers is beyond the scope of this work, but are worth further study.
+Explaining such outliers is beyond the scope of this work, but these targets are worth further study as they may reveal new physics.
 
@@
 Given the systematic offset of $\sim$0.5 mag, a fairly wide band of stars ($0\le \Delta M_G \le 1$) was selected as being ``close to the main sequence''.
-This final sample included 440 stars.
+The final sample included 440 stars.
 This simplistic cut is not a robust dwarf--giant, nor single--binary star separator, but serves to select a sample of mostly main sequence stars for the illustrative purpose of this work.
@@
 
-%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%
 \section{Extending the Spin-Down Gap}
+A bimodal period distribution was first discovered by \citet{mcquillan2013} for \Kepler M dwarfs, who found a dearth of objects with periods around $\sim$25 days.
+Follow-up work by \citet{mcquillan2014} found this bimodality extended to K dwarfs, up to $T_{eff}\sim5500$.
+Figure \ref{fig:gyro} shows the rotation period distribution for the final 440 star sample of likely main sequence stars.
+The bimodality appears to extend smoothly through to the hottest stars in this sample.
 
-this feature first observed in rotation period by \citet{mcquillan2013} for M dwarfs
-
-here we show it appears to extended to nearby higher mass stars in the \Kepler field.
-
-the reason this feature did not appear in earlier \Kepler studies is the inability to select main sequence stars from turn-off or subgiants using the KIC
+While these periods for field stars were robustly measured by \citet{mcquillan2014} and others, the bimodality did not appear in previous \Kepler work due to the high contamination rate by subgiants for these bluer, hotter stars.
+414 rotating stars had acceptable photometric and parallax uncertainties in TGAS, but were culled from this sample for having $M_G$ luminosities higher than the main sequence cut in \S3 above.
+Note that distributions of $\log g$ values from the Kepler Input Catalog \citep{brown2011a} for both the main sequence and subgiant stars were not statistically different.
 
@@
 \caption{ Rotation period versus temperature for TGAS-matched stars near the isochrone main sequence (blue circles). 
-%Distributions of rotation periods for these main sequence selected stars are shown as vertical histograms for 7 bins of temperature (black lines), with each temperature bin having equal numbers of stars.
-As in Figure \ref{fig:all}, the full \Kepler--Gaia matched sample is shown for reference (small black dots).
-The bimodality in rotation periods previously seen by \citet{mcquillan2014} extends the full range of temperatures in the \Kepler--Gaia main sequence sample shown here.
-A \citet{meibom2011} 600 Myr gyrochonology-isochrone roughly traces the bimodality midpoint up to 6000 K (red solid line), but deviates sharply up to $\sim$6200 K (red dotted line).
-A log-linear extrapolation to of the isochrone from 6000K to 6500 K (red dashed line) continues to track the bimodality to hotter temperatures, and roughly traces a line of constant Rossby number.
+The full \Kepler--Gaia matched sample is shown for reference (small black dots).
+A bimodality in rotation periods initially discovered for M dwarfs by \citet{mcquillan2013}, extends the full range of temperatures in the \Kepler--Gaia main sequence sample shown here.
+A \citet{meibom2011} 600 Myr gyrochonology-isochrone (gyrochrone) traces the bimodality midpoint up to 6000 K (red solid line), but deviates from the isochrone sharply to $\sim$6200 K (red dotted line).
+A log-linear extrapolation of the isochrone at 6000K to 6500 K (red dashed line) continues to track the bimodality to hotter temperatures, and roughly follows a line of constant Rossby number.
 }
@@
 
+The minimum in the bimodal period distribution in Figure \ref{fig:gyro} can be traced using a gyrochonology isochrone (colloquially known as a ``gyrochrone'').
+Many different studies have produced competing gyrochrone models, each with unique morphologies at the hot and cool star regimes \citep[e.g.][]{barnes2007,mamajek2008,meibom2011,angus2015}.
+A 600 Myr gyrochrone from \citet{meibom2011} was determined by eye to approximately trace the period minima from 3500 K to 6000 K.
+As shown in Figure \ref{fig:gyro}, this model (as with most gyrochronology models) turns down in period sharply for stars hotter than $\sim$6000 K.
+A log-linear extrapolation of the 600 Myr gyrochrone at 6000 K continues to trace the period bimodality up to 6500 K, and roughly traces a line of constant Rossby number.
+
+%%%%%
 \begin{figure}[]
@@
 \includegraphics[width=4in]{fig5.png}
-\caption{Residual of log rotation periods about the \citet{meibom2011} 600 Myr isochrone, using the log-linear extrapolation between 6000--6500 K shown in Figure \ref{fig:gyro}.
+\caption{Residual of log rotation periods about the \citet{meibom2011} 600 Myr gyrochrone, using the log-linear extrapolation between 6000--6500 K shown in Figure \ref{fig:gyro}.
 The bimodal rotation period distribution is clear, with peaks at $\Delta \log P_{rot}$ of -0.19 and 0.21 days.
 A two-Gaussian model was fit to this distribution (blue line).
-}
+Approximately 50\% more stars are present in the slower, nominally older peak (right).}
 \label{fig:diff}
 \end{figure}
+
+The difference (in log period) between the observed rotation and the 600 Myr gyrochrone for the 440 likely main sequence stars is shown in Figure \ref{fig:diff}.
+Despite combining stars of all temperatures, the bimodality is clearly seen in this log period space.
+A two Gaussian model was fit to this data, which found peaks in the two periods distributions of $-0.19 \pm 0.01$ and $0.21 \pm 0.01$ days.
+262 stars had periods longer than the gyrochrone model (right peak) and 178 slower than the model (left peak).
+This is in contrast to the overall results from \citet{mcquillan2013} who found nearly equal numbers of M dwarfs in the fast and slow rotating groups, but is in general agreement with their sample with non-zero proper motions.
+%IDL> print, 262./178.
+%      1.47191 ==> US
+%IDL> print, 345./266.
+%      1.29699 ==> THEM
+
+%%%%%
+\begin{figure}[]
+\centering
+\includegraphics[width=4in]{fig6.png}
+\caption{Total proper motion distributions for stars above (rotating slower, older stars) the gyrochrone model shown in Figure \ref{fig:gyro} (black) and below (rotating faster, younger stars) the model (green).
+Median values of the two distributions are shown (thick lines), which yield a marginal 2.8$\sigma$ difference.
+}
+\label{fig:pm}
+\end{figure}
+
+this age explanation was bolstered in \citet{mcquillan2013}, who noted the two populations of rotating M dwarfs had different distributions of proper motions, indicating they belonged to distinct groups of stars.
+The distribution of total proper motion for stars above and below the modified 600 Myr gyrochrone is shown in Figure \ref{fig:pm}.
+Stars above the gyrochrone (slower rotators, nominally older) have a median total proper motion of 15.4 mas/yr, while those below (faster rotators, younger)  have a median of 11.3 mas/yr.
+This difference in kinematics versus rotation period is in the same direction observed by \citet{mcquillan2013}.
+As the typical error in the total proper motion is $\sim$1.4 mas/yr for this sample, this is a marginal (2.8$\sigma$) difference.
+Note also that likely contamination of subgiants in the hotter stars may also obscure kinematic differences.
 
@@
 this similar to the non-continusous star formation history for the local solar neighborhood using Hipparcos astrometry \citep{hernandez2000}.
-this age explanation was bolstered in \citet{mcquillan2013}, who noted the two populations of rotating M dwarfs had different distributions of proper motions, indicating they belonged to distinct groups of stars.
-The distribution of total proper motion for stars above and below the modified 600 Myr gyrochrone is shown in Figure \ref{fig:pm}.
-Stars above the gyrochrone (slower rotators, nominally older) have a median total proper motion of 15.4 mas/yr, while those below (faster rotators, younger)  have a median of 11.3 mas/yr.
-This difference in kinematics versus rotation period is in the same direction observed by \citet{mcquillan2013}.
-As the typical error in the total proper motion is $\sim$1.4 mas/yr for this sample, this is a marginal (2.8$\sigma$) difference.
-Note also that likely contamination of subgiants in the hotter stars may also obscure kinematic differences.
-
-%('median above: ', 15.373856668858519)
-%('median below: ', 11.301950163653459)
-%('median PM error:', 1.4275539188520987)
-%('diff / error = ', 2.8523661708548951)
-\begin{figure}[]
-\centering
-\includegraphics[width=4in]{fig6.png}
-\caption{Total proper motion distributions for stars above (rotating slower, older stars) the gyrochrone model shown in Figure \ref{fig:gyro} (black) and below (rotating faster, younger stars) the model (green).
-Median values of the two distributions are shown (thick lines), which yield a marginal 2.8$\sigma$ difference.
-}
-\label{fig:pm}
-\end{figure}
 

### 2016-10-21 paper/ms.tex
full draft finised. win

@@
 These main sequence stars show a bimodality in their rotation period distribution, centered roughly around a 600 Myr rotation-isochrone.
-This feature matches the bimodal period distribution from cooler stars with \Kepler, but was previously undetected for solar-type stars due to sample contamination by subgiants.
+This feature matches the bimodal period distribution found in cooler stars with \Kepler, but was previously undetected for solar-type stars due to sample contamination by subgiants.
 A tenuous connection between the rotation period and total proper motion is found, suggesting the period bimodality is due to the age distribution of stars within $\sim$300pc of the Sun, rather than a phase of rapid angular momentum loss.
-This emphasizes the unique power for stellar populations studies of combining temporal monitoring from \Kepler with astrometric data from Gaia
+This work emphasizes the unique power for understanding stellar populations by combining temporal monitoring from \Kepler with astrometric data from Gaia.
 \end{abstract}
@@
 As shown in Figure \ref{fig:gyro}, this model (as with most gyrochronology models) turns down in period sharply for stars hotter than $\sim$6000 K.
-A log-linear extrapolation of the 600 Myr gyrochrone at 6000 K continues to trace the period bimodality up to 6500 K, and roughly traces a line of constant Rossby number.
+A log-linear extrapolation of the 600 Myr gyrochrone at 6000 K continues to trace the period bimodality up to 6500 K, and roughly traces a line of constant Rossby number assuming a local convective turnover timescale such as from \citet{barnes_kim2010}.
 
@@
 Despite combining stars of all temperatures, the bimodality is clearly seen in this log period space.
-A two Gaussian model was fit to this data, which found peaks in the two periods distributions of $-0.19 \pm 0.01$ and $0.21 \pm 0.01$ days.
+A two Gaussian model was fit to this data, which found peaks in the two distributions of $-0.19 \pm 0.01$ and $0.21 \pm 0.01$ days.
 262 stars had periods longer than the gyrochrone model (right peak) and 178 slower than the model (left peak).
-This is in contrast to the overall results from \citet{mcquillan2013} who found nearly equal numbers of M dwarfs in the fast and slow rotating groups, but is in general agreement with their sample with non-zero proper motions.
+This is in contrast to the overall results from \citet{mcquillan2013} who found nearly equal numbers of M dwarfs in the fast and slow rotating groups, but is in general agreement with their sample of stars with non-zero proper motions.
 %IDL> print, 262./178.
@@
 
-this age explanation was bolstered in \citet{mcquillan2013}, who noted the two populations of rotating M dwarfs had different distributions of proper motions, indicating they belonged to distinct groups of stars.
-The distribution of total proper motion for stars above and below the modified 600 Myr gyrochrone is shown in Figure \ref{fig:pm}.
+The favored explanation for the bimodal period distribution by \citet{mcquillan2013} was an age effect, with nearby stars having a bimodal star formation history.
+This explanation was bolstered by their observation that stars in the two period groups have differing distributions of proper motions, indicating they belonged to kinematically separate groups.
+This measurement is replicated in Figure \ref{fig:pm}, which shows the total proper motion distribution for stars above and below the 600 Myr gyrochrone.
 Stars above the gyrochrone (slower rotators, nominally older) have a median total proper motion of 15.4 mas/yr, while those below (faster rotators, younger)  have a median of 11.3 mas/yr.
 This difference in kinematics versus rotation period is in the same direction observed by \citet{mcquillan2013}.
-As the typical error in the total proper motion is $\sim$1.4 mas/yr for this sample, this is a marginal (2.8$\sigma$) difference.
-Note also that likely contamination of subgiants in the hotter stars may also obscure kinematic differences.
+As the typical error in the  Gaia DR1 total proper motion is $\sim$1.4 mas/yr for this sample, this is a statistically marginal (2.8$\sigma$) difference.
 
@@
 
-median distance of K/M stars ($T_{eff} > 4000$ K) is $\sim$216 pc, using isochrone
-median distance for the TGAS-matched sample (blue points) is very close, 285 pc, so also sampling just the most local volume.
-this points to the effect being localized around us, which explains why it has not been seen in other gyro studies to date (e.g. clusters)
-Unfortunately not enough rotation period data across the fully convective boundary ($\sim$3000 K) to tell if this bimodal feature continues.
+Using a combination of data from \Kepler and Gaia DR1, I have explored the rotation period distribution for 894 nearby main sequence stars.
+A bimodal rotation period distribution has been found in stars with temperatures ranging from 5000 K to 6500 K.
+This feature matches that found in cooler stars from \Kepler, but was only revealed thanks to the enhanced ability to distinguish dwarfs from subgiants using Gaia data.
+A tenuous difference in the TGAS total proper motion for stars in the fast and slow rotating groups is found, which is in agreement with the findings for cool stars by \citet{mcquillan2013}.
 
-the bimodality may be another manifestation of the ``Vaughan Preston gap'' \citep{vaughan1980}, e.g. discussed for rotating stars by\citet{kado-fong2016}.
-either due to fast evolution through intermediate stellar activity, or could be an age gap.
-if age, these clumps line up roughly with a 300 Myr and 2 Gyr \citet{meibom2011} gyro-isochrone from cooler than about 6000 K.
-this similar to the non-continusous star formation history for the local solar neighborhood using Hipparcos astrometry \citep{hernandez2000}.
+While a definitive explanation for this period bimodality has not been reached, the findings to date seem to favor stellar ages as the cause.
+In this scenario the star formation history for nearby stars would be dominated by two main bouts of star formation, one centered at a few hundred Myr, and one at a few Gyr (slightly younger than the Sun).
+It is also worth noting that the space volume probed by the TGAS sample investigated here is very similar to that covered by the temperature-selected cool star sample in \citet{mcquillan2013}.
+The median parallax distance for stars in this work is 285 pc, while the median isochrone distance for the K and M dwarfs is $\sim$216 pc.
+This points to the period distribution being a localized age artifact, and explains why previous exhaustive studies of rotation periods in stellar clusters did not find such a bimodality.
+Determining how localized this age distribution is, and if it can be confirmed for stars across the HR diagram including giants, is a key goal for future Gaia data releases.
 
-finally, %gaia + \Kepler useful for determining the age distribution of the nearby field
-have shown utility for using Gaia data, combined with detailed light curve statitiscs from \Kepler, to reveal hidden structure in properties of field stars.
-This combination will be super useful for determining approximate log g values (separating main sequence from sub-giants), and thus more accurate gyrochronology ages \citep{van-saders2013}.
-also for modeling the star formation history of the entire milky way \citep[e.g.][]{bertelli1999}.
+The period bimodality may yet be a manifestation of the ``Vaughan-Preston'' gap observed in chromospheric activity indicators from solar type stars.
+Such a feature has also been discussed for rotating stars by\citet{kado-fong2016}.
+Given that the mass range for the bimodality explored here and in \citet{mcquillan2014} covers stars with solar-type dynamos (those having a tachocline, late F through early M) such a model cannot be fully ruled out at this time.
+Unfortunately too few rotation periods have been measured for stars across the ``fully convective boundary'' ($T_{eff}<3000$ K, spectral type $\sim$M4) to tell if the bimodal period feature continues to cooler temperatures, which would support the age distribution model.
+
+Finally, this exploratory work has highlighted the utility of using astrometric data from Gaia combined with detailed light curve statistics from \Kepler to reveal hidden substructure in the properties of field stars.
+Looking forward to the astrometric precision of future Gaia data releases, this combination will be effective at separating dwarf stars from subgiants for nearly the entire \Kepler and K2 databases, and enable accurate age maps for field stars.
+% super useful for determining approximate log g values (separating main sequence from sub-giants), and thus more accurate gyrochronology ages \citep{van-saders2013}.
 
@@
 \acknowledgments
-Special thanks to Jennifer van Saders, Sean Matt, and Travis Metcalfe for their helpful discussions that motivated publishing this work.
-
+Special thanks to Kevin Covey, Jennifer van Saders, Sean Matt, and Travis Metcalfe for their helpful discussions that motivated publishing this work.
 JRAD is supported by an NSF Astronomy and Astrophysics Postdoctoral Fellowship under award AST-1501418.
@@
 
-\bibliography{/Users/davenpj3/Dropbox/references}
+\bibliography{/Users/james/Dropbox/references}
 

### 2016-10-21 paper/ms.tex
add a few more references

@@
 \centering
-\includegraphics[width=5.3in]{fig4alt.png}
+\includegraphics[width=5.3in]{fig4.png}
 \caption{ Rotation period versus temperature for TGAS-matched stars near the isochrone main sequence (blue circles). 
@@
 Given that the mass range for the bimodality explored here and in \citet{mcquillan2014} covers stars with solar-type dynamos (those having a tachocline, late F through early M) such a model cannot be fully ruled out at this time.
-Unfortunately too few rotation periods have been measured for stars across the ``fully convective boundary'' ($T_{eff}<3000$ K, spectral type $\sim$M4) to tell if the bimodal period feature continues to cooler temperatures, which would support the age distribution model.
+Though there have been many rotation studies for cool stars \citep[e.g.][]{irwin2011,newton2016,stelzer2016} too few rotation periods have been measured for stars across the ``fully convective boundary'' ($T_{eff}<3000$ K, spectral type $\sim$M4) to tell if the bimodal period feature continues to cooler temperatures, which would support the age distribution model.
+Further studies of rotation periods for stars in intermediate age open clusters (e.g.
+the Hyades) may also help solve this mystery \citep[e.g.][]{douglas2014}.
 
@@
 \acknowledgments
-Special thanks to Kevin Covey, Jennifer van Saders, Sean Matt, and Travis Metcalfe for their helpful discussions that motivated publishing this work.
+Special thanks to Kevin Covey, Jennifer van Saders, Sean Matt, Travis Metcalfe, and Aleks Scholz for their helpful discussions that motivated publishing this work.
 JRAD is supported by an NSF Astronomy and Astrophysics Postdoctoral Fellowship under award AST-1501418.

### 2016-10-24 paper/ms.tex
change to PDF
add acknl for GaiaSprint and Gaia

@@
 \centering
-\includegraphics[width=6in]{fig1.png}
+\includegraphics[width=6in]{fig1}
 \caption{
@@
 \centering
-\includegraphics[width=5.3in]{fig2.png}
+\includegraphics[width=5.3in]{fig2}
 \caption{Hertzsprung--Russell (HR) diagram using temperatures from \citet{mcquillan2014} and Gaia DR1 $G$-band absolute magnitudes for the 894 stars that pass photometric and parallax quality cuts described in the text.
@@
 \centering
-\includegraphics[width=4in]{fig3.png}
+\includegraphics[width=4in]{fig3}
 \caption{Distribution of differences between the Gaia DR1 $G$ absolute magnitudes and values for main sequence stars from the 300 Myr isochrone shown in Figure \ref{fig:HR}.
@@
 \centering
-\includegraphics[width=5.3in]{fig4.png}
+\includegraphics[width=5.3in]{fig4}
 \caption{ Rotation period versus temperature for TGAS-matched stars near the isochrone main sequence (blue circles). 
@@
 \centering
-\includegraphics[width=4in]{fig5.png}
+\includegraphics[width=4in]{fig5}
 \caption{Residual of log rotation periods about the \citet{meibom2011} 600 Myr gyrochrone, using the log-linear extrapolation between 6000--6500 K shown in Figure \ref{fig:gyro}.
@@
 \centering
-\includegraphics[width=4in]{fig6.png}
+\includegraphics[width=4in]{fig6}
 \caption{Total proper motion distributions for stars above (rotating slower, older stars) the gyrochrone model shown in Figure \ref{fig:gyro} (black) and below (rotating faster, younger stars) the model (green).
@@
 
+This project was developed as part of the 2016 NYC Gaia Sprint, hosted by the Center for Computational Astrophysics at the Simons Foundation in New York City.
+
 This research made use of the cross-match service provided by CDS, Strasbourg.
 
-\bibliography{/Users/james/Dropbox/references}
+This work has made use of data from the European Space Agency (ESA)
+mission {\it Gaia} (\url{http://www.cosmos.esa.int/gaia}), processed by
+the {\it Gaia} Data Processing and Analysis Consortium (DPAC,
+\url{http://www.cosmos.esa.int/web/gaia/dpac/consortium}). Funding
+for the DPAC has been provided by national institutions, in particular
+the institutions participating in the {\it Gaia} Multilateral Agreement.
+
+\begin{thebibliography}{}
+\expandafter\ifx\csname natexlab\endcsname\relax\def\natexlab#1{#1}\fi
+
+\bibitem[{{Aigrain} {et~al.}(2015){Aigrain}, {Llama}, {Ceillier}, {Chagas},
+  {Davenport}, {Garc{\'{\i}}a}, {Hay}, {Lanza}, {McQuillan}, {Mazeh}, {de
+  Medeiros}, {Nielsen}, \& {Reinhold}}]{aigrain2015}
+{Aigrain}, S., {Llama}, J., {Ceillier}, T., {et~al.} 2015, \mnras, 450, 3211
+
+\bibitem[{{Angus} {et~al.}(2015){Angus}, {Aigrain}, {Foreman-Mackey}, \&
+  {McQuillan}}]{angus2015}
+{Angus}, R., {Aigrain}, S., {Foreman-Mackey}, D., \& {McQuillan}, A. 2015,
+  \mnras, 450, 1787
+
+\bibitem[{{Barnes}(2007)}]{barnes2007}
+{Barnes}, S.~A. 2007, \apj, 669, 1167
+
+\bibitem[{{Barnes} \& {Kim}(2010)}]{barnes_kim2010}
+{Barnes}, S.~A., \& {Kim}, Y.-C. 2010, \apj, 721, 675
+
+\bibitem[{{Bertelli} {et~al.}(1999){Bertelli}, {Bressan}, {Chiosi}, \&
+  {Vallenari}}]{bertelli1999}
+{Bertelli}, G., {Bressan}, A., {Chiosi}, C., \& {Vallenari}, A. 1999, Baltic
+  Astronomy, 8, 271
+
+\bibitem[{{Borucki} {et~al.}(2010){Borucki}, {Koch}, {Basri}, {Batalha},
+  {Brown}, {Caldwell}, {Caldwell}, {Christensen-Dalsgaard}, {Cochran},
+  {DeVore}, {Dunham}, {Dupree}, {Gautier}, {Geary}, {Gilliland}, {Gould},
+  {Howell}, {Jenkins}, {Kondo}, {Latham}, {Marcy}, {Meibom}, {Kjeldsen},
+  {Lissauer}, {Monet}, {Morrison}, {Sasselov}, {Tarter}, {Boss}, {Brownlee},
+  {Owen}, {Buzasi}, {Charbonneau}, {Doyle}, {Fortney}, {Ford}, {Holman},
+  {Seager}, {Steffen}, {Welsh}, {Rowe}, {Anderson}, {Buchhave}, {Ciardi},
+  {Walkowicz}, {Sherry}, {Horch}, {Isaacson}, {Everett}, {Fischer}, {Torres},
+  {Johnson}, {Endl}, {MacQueen}, {Bryson}, {Dotson}, {Haas}, {Kolodziejczak},
+  {Van Cleve}, {Chandrasekaran}, {Twicken}, {Quintana}, {Clarke}, {Allen},
+  {Li}, {Wu}, {Tenenbaum}, {Verner}, {Bruhweiler}, {Barnes}, \&
+  {Prsa}}]{borucki2010}
+{Borucki}, W.~J., {Koch}, D., {Basri}, G., {et~al.} 2010, Science, 327, 977
+
+\bibitem[{{Bressan} {et~al.}(2012){Bressan}, {Marigo}, {Girardi}, {Salasnich},
+  {Dal Cero}, {Rubele}, \& {Nanni}}]{bressan2012}
+{Bressan}, A., {Marigo}, P., {Girardi}, L., {et~al.} 2012, \mnras, 427, 127
+
+\bibitem[{{Brown} {et~al.}(2011){Brown}, {Latham}, {Everett}, \&
+  {Esquerdo}}]{brown2011a}
+{Brown}, T.~M., {Latham}, D.~W., {Everett}, M.~E., \& {Esquerdo}, G.~A. 2011,
+  \aj, 142, 112
+
+\bibitem[{{Carrasco} {et~al.}(2014){Carrasco}, {Catal{\'a}n}, {Jordi},
+  {Tremblay}, {Napiwotzki}, {Luri}, {Robin}, \& {Kowalski}}]{carrasco2014}
+{Carrasco}, J.~M., {Catal{\'a}n}, S., {Jordi}, C., {et~al.} 2014, \aap, 565,
+  A11
+
+\bibitem[{{Ciardi} {et~al.}(2011){Ciardi}, {von Braun}, {Bryden}, {van Eyken},
+  {Howell}, {Kane}, {Plavchan}, {Ram{\'{\i}}rez}, \& {Stauffer}}]{ciardi2011}
+{Ciardi}, D.~R., {von Braun}, K., {Bryden}, G., {et~al.} 2011, \aj, 141, 108
+
+\bibitem[{{Creevey} {et~al.}(2013){Creevey}, {Th{\'e}venin}, {Basu}, {Chaplin},
+  {Bigot}, {Elsworth}, {Huber}, {Monteiro}, \& {Serenelli}}]{creevey2013}
+{Creevey}, O.~L., {Th{\'e}venin}, F., {Basu}, S., {et~al.} 2013, \mnras, 431,
+  2419
+
+\bibitem[{{do Nascimento} {et~al.}(2012){do Nascimento}, {da Costa}, \&
+  {Castro}}]{donascimento2012}
+{do Nascimento}, J.-D., {da Costa}, J.~S., \& {Castro}, M. 2012, \aap, 548, L1
+
+\bibitem[{{Douglas} {et~al.}(2014){Douglas}, {Ag{\"u}eros}, {Covey}, {Bowsher},
+  {Bochanski}, {Cargile}, {Kraus}, {Law}, {Lemonias}, {Arce}, {Fierroz}, \&
+  {Kundert}}]{douglas2014}
+{Douglas}, S.~T., {Ag{\"u}eros}, M.~A., {Covey}, K.~R., {et~al.} 2014, \apj,
+  795, 161
+
+\bibitem[{{Gaensicke} {et~al.}(2015){Gaensicke}, {Tremblay}, {Barstow}, {Bono},
+  {Burleigh}, {Casewell}, {Dhillon}, {Farihi}, {Garcia-Berro}, {Geier},
+  {Gentile-Fusillo}, {Hermes}, {Hollands}, {Istrate}, {Jordan}, {Knigge},
+  {Manser}, {Marsh}, {Nelemans}, {Pala}, {Raddi}, {Tauris}, {Toloza}, {Veras},
+  {Werner}, \& {Wilson}}]{gaensicke2015}
+{Gaensicke}, B., {Tremblay}, P.-E., {Barstow}, M., {et~al.} 2015, ArXiv
+  e-prints, arXiv:1506.02653
+
+\bibitem[{{Gaia Collaboration}(2016)}]{gaia}
+{Gaia Collaboration}. 2016, ArXiv e-prints, arXiv:1609.04153
+
+\bibitem[{{Hernandez} {et~al.}(2000){Hernandez}, {Valls-Gabaud}, \&
+  {Gilmore}}]{hernandez2000}
+{Hernandez}, X., {Valls-Gabaud}, D., \& {Gilmore}, G. 2000, \mnras, 316, 605
+
+\bibitem[{{Irwin} {et~al.}(2011){Irwin}, {Berta}, {Burke}, {Charbonneau},
+  {Nutzman}, {West}, \& {Falco}}]{irwin2011}
+{Irwin}, J., {Berta}, Z.~K., {Burke}, C.~J., {et~al.} 2011, \apj, 727, 56
+
+\bibitem[{{Kado-Fong} {et~al.}(2016){Kado-Fong}, {Williams}, {Mann}, {Berger},
+  {Burgett}, {Chambers}, {Huber}, {Kaiser}, {Kudritzki}, {Magnier},
+  {Wainscoat}, \& {Waters}}]{kado-fong2016}
+{Kado-Fong}, E., {Williams}, P.~K.~G., {Mann}, A.~W., {et~al.} 2016, ArXiv
+  e-prints, arXiv:1608.00978
+
+\bibitem[{{Lindegren} {et~al.}(2016){Lindegren}, {Lammers}, {Bastian},
+  {Hern{\'a}ndez}, {Klioner}, {Hobbs}, {Bombrun}, {Michalik}, {Ramos-Lerate},
+  {Butkevich}, {Comoretto}, {Joliet}, {Holl}, {Hutton}, {Parsons},
+  {Steidelm{\"u}ller}, {Abbas}, {Altmann}, {Andrei}, {Anton}, {Bach},
+  {Barache}, {Becciani}, {Berthier}, {Bianchi}, {Biermann}, {Bouquillon},
+  {Bourda}, {Br{\"u}semeister}, {Bucciarelli}, {Busonero}, {Carlucci},
+  {Casta{\~n}eda}, {Charlot}, {Clotet}, {Crosta}, {Davidson}, {de Felice},
+  {Drimmel}, {Fabricius}, {Fienga}, {Figueras}, {Fraile}, {Gai}, {Garralda},
+  {Geyer}, {Gonz{\'a}lez-Vidal}, {Guerra}, {Hambly}, {Hauser}, {Jordan},
+  {Lattanzi}, {Lenhardt}, {Liao}, {L{\"o}ffler}, {McMillan}, {Mignard}, {Mora},
+  {Morbidelli}, {Portell}, {Riva}, {Sarasso}, {Serraller}, {Siddiqui}, {Smart},
+  {Spagna}, {Stampa}, {Steele}, {Taris}, {Torra}, {van Reeven}, {Vecchiato},
+  {Zschocke}, {de Bruijne}, {Gracia}, {Raison}, {Lister}, {Marchant},
+  {Messineo}, {Soffel}, {Osorio}, {de Torres}, \& {O'Mullane}}]{gaia_dr1}
+{Lindegren}, L., {Lammers}, U., {Bastian}, U., {et~al.} 2016, ArXiv e-prints,
+  arXiv:1609.04303
+
+\bibitem[{{Lurie} {et~al.}(2015){Lurie}, {Davenport}, {Hawley}, {Wilkinson},
+  {Wisniewski}, {Kowalski}, \& {Hebb}}]{lurie2015}
+{Lurie}, J.~C., {Davenport}, J.~R.~A., {Hawley}, S.~L., {et~al.} 2015, \apj,
+  800, 95
+
+\bibitem[{{Mamajek} \& {Hillenbrand}(2008)}]{mamajek2008}
+{Mamajek}, E.~E., \& {Hillenbrand}, L.~A. 2008, \apj, 687, 1264
+
+\bibitem[{{Mann} {et~al.}(2012){Mann}, {Gaidos}, {L{\'e}pine}, \&
+  {Hilton}}]{mann2012}
+{Mann}, A.~W., {Gaidos}, E., {L{\'e}pine}, S., \& {Hilton}, E.~J. 2012, \apj,
+  753, 90
+
+\bibitem[{{Massarotti}(2008)}]{massarotti2008}
+{Massarotti}, A. 2008, \aj, 135, 2287
+
+\bibitem[{{McQuillan} {et~al.}(2013){McQuillan}, {Aigrain}, \&
+  {Mazeh}}]{mcquillan2013}
+{McQuillan}, A., {Aigrain}, S., \& {Mazeh}, T. 2013, \mnras, 432, 1203
+
+\bibitem[{{McQuillan} {et~al.}(2014){McQuillan}, {Mazeh}, \&
+  {Aigrain}}]{mcquillan2014}
+{McQuillan}, A., {Mazeh}, T., \& {Aigrain}, S. 2014, \apjs, 211, 24
+
+\bibitem[{{Meibom} {et~al.}(2011){Meibom}, {Barnes}, {Latham}, {Batalha},
+  {Borucki}, {Koch}, {Basri}, {Walkowicz}, {Janes}, {Jenkins}, {Van Cleve},
+  {Haas}, {Bryson}, {Dupree}, {Furesz}, {Szentgyorgyi}, {Buchhave}, {Clarke},
+  {Twicken}, \& {Quintana}}]{meibom2011}
+{Meibom}, S., {Barnes}, S.~A., {Latham}, D.~W., {et~al.} 2011, \apjl, 733, L9
+
+\bibitem[{{Newton} {et~al.}(2016){Newton}, {Irwin}, {Charbonneau},
+  {Berta-Thompson}, {Dittmann}, \& {West}}]{newton2016}
+{Newton}, E.~R., {Irwin}, J., {Charbonneau}, D., {et~al.} 2016, \apj, 821, 93
+
+\bibitem[{{Reinhold} {et~al.}(2013){Reinhold}, {Reiners}, \&
+  {Basri}}]{reinhold2013}
+{Reinhold}, T., {Reiners}, A., \& {Basri}, G. 2013, \aap, 560, A4
+
+\bibitem[{{Skumanich}(1972)}]{skumanich1972}
+{Skumanich}, A. 1972, \apj, 171, 565
+
+\bibitem[{{Stassun} \& {Torres}(2016)}]{stassun2016}
+{Stassun}, K.~G., \& {Torres}, G. 2016, ArXiv e-prints, arXiv:1609.05390
+
+\bibitem[{{Stelzer} {et~al.}(2016){Stelzer}, {Damasso}, {Scholz}, \&
+  {Matt}}]{stelzer2016}
+{Stelzer}, B., {Damasso}, M., {Scholz}, A., \& {Matt}, S.~P. 2016, \mnras, 463,
+  1844
+
+\bibitem[{{Tayar} {et~al.}(2015){Tayar}, {Ceillier},
+  {Garc{\'{\i}}a-Hern{\'a}ndez}, {Troup}, {Mathur}, {Garc{\'{\i}}a}, {Zamora},
+  {Johnson}, {Pinsonneault}, {M{\'e}sz{\'a}ros}, {Allende Prieto}, {Chaplin},
+  {Elsworth}, {Hekker}, {Nidever}, {Salabert}, {Schneider}, {Serenelli},
+  {Shetrone}, \& {Stello}}]{tayar2015}
+{Tayar}, J., {Ceillier}, T., {Garc{\'{\i}}a-Hern{\'a}ndez}, D.~A., {et~al.}
+  2015, \apj, 807, 82
+
+\bibitem[{{van Saders} {et~al.}(2016){van Saders}, {Ceillier}, {Metcalfe},
+  {Silva Aguirre}, {Pinsonneault}, {Garc{\'{\i}}a}, {Mathur}, \&
+  {Davies}}]{van-saders2016}
+{van Saders}, J.~L., {Ceillier}, T., {Metcalfe}, T.~S., {et~al.} 2016, \nat,
+  529, 181
+
+\bibitem[{{Vaughan} \& {Preston}(1980)}]{vaughan1980}
+{Vaughan}, A.~H., \& {Preston}, G.~W. 1980, \pasp, 92, 385
+
+\end{thebibliography}
 

### 2016-10-24 paper/ms.tex
update more acknl

@@
 The Tycho-Gaia Astrometric Solution (TGAS) measures improved proper motions and parallaxes for 2 million nearby, bright sources by extending the astrometric solutions from Tycho and Hipparcos.
-While the TGAS data are not a complete astrometric survey, and have possible systematics in the reported parallaxes \citep{stassun2016}, they represent a significant improvement in the astrometry and kinematics available for stars in the \Kepler field thanks to the precision of Gaia.
+While the TGAS data are not a complete astrometric survey, and have possible systematics in the reported parallaxes \citep{gaia_dr1,stassun2016}, they represent a significant improvement in the astrometry and kinematics available for stars in the \Kepler field thanks to the precision of Gaia.
 
@@
 Due to the brightness limits of the TGAS sample very few K and M dwarfs were recovered in the TGAS sample.
-Future releases of Gaia data will ostensibly provide full astrometric solutions for nearly all \Kepler stars.
+Future releases of Gaia data will provide full astrometric solutions for nearly all \Kepler stars.
 The rotation periods versus stellar effective temperatures for the \Kepler--Gaia matched stars are shown in Figure \ref{fig:all}.
@@
 Faint stars were removed by requiring sources have $G$-band flux errors $<1$\%.
-To ensure accurate distances, and therefore luminosities, parallaxes were required to have errors $<0.4$ mas year$^{-1}$.
+To ensure accurate distances, and therefore luminosities, parallaxes were required to have errors $<0.4$ mas.
 These cuts left a total of 894 stars from the \Kepler--TGAS matched sample (68\%).
@@
 \bibitem[{{Gaia Collaboration}(2016)}]{gaia}
-{Gaia Collaboration}. 2016, ArXiv e-prints, arXiv:1609.04153
+{Gaia Collaboration}. 2016, ArXiv e-prints, arXiv:1609.04153, doi:10.1051/0004-6361/201629272
 

### 2016-10-26 paper/ms.tex
paper submitted

@@
 At present, ages determined by gyrochronology are accurate to $\sim$10\% in the {\it best} cases (young solar-type stars).
-With rotation now being routinely measured, it is hoped a new ability of determining robust ages for field stars will be possible by calibrating gyro-isochrones to stellar clusters and asteroseismic samples, and improved models of angular momentum loss \citep[e.g.][]{angus2015,van-saders2016}.
+Determining robust ages for field stars may soon be possible by calibrating gyro-isochrones to stellar clusters and asteroseismic samples, and improved models of angular momentum loss \citep[e.g.][]{angus2015,van-saders2016}.
 
-Curiously, in studying the rotation periods for M dwarfs in the \Kepler field, \citet{mcquillan2013} discovered a bimodal period distribution, which was subsequently confirmed to exist up to K dwarfs in the \Kepler field by \citet{mcquillan2014}.
-However, this had never been observed in any other study of stellar rotation periods, including stellar clusters at a variety of ages, nor was it detected in the \Kepler stars at hotter temperatures ($T_{eff} > 5000$).
-While binary stars and multiple-period systems may be contaminating the rotation period sample for \Kepler field stars, \citet{mcquillan2014} found they could not adequately explain the bimodal period distribution.
+\citet{mcquillan2013} discovered a bimodal period distribution for M dwarfs in the \Kepler field, which was subsequently confirmed to exist for K dwarfs as well \citep{mcquillan2014}.
+However, this bimodality had never been observed in any other study of stellar rotation periods, including stellar clusters at a variety of ages, nor was it detected in the \Kepler stars at hotter temperatures ($T_{eff} > 5000$).
+While binary stars and multiple-period systems may be contaminating the rotation period sample for \Kepler field stars, \citet{mcquillan2014} found the presence of such interlopers could not adequately explain the bimodal period distribution.
 Currently favored explanations for this feature are 1) a non-continuous age distribution for nearby stars, as was suggested with very nearby Hipparcos stars by \citet{hernandez2000}, or 2) a previously unknown phase of rapid angular momentum loss for low-mass stars, similar to the ``Vaughan-Preston'' gap seen in chromospheric activity indicators \citep{vaughan1980}.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%
-\section{The \Kepler--Gaia Data}
-Rotation periods in this study come from \citet{mcquillan2014}, who performed an Auto-Correlation Function analysis of \Kepler stars cooler than 6500 K that had at least $\sim$2 years of observation.
-The periods recovered from this approach generally agree very well with those found via Lomb-Scargle Periodograms \citep[e.g.][]{reinhold2013,aigrain2015}.
-Sources with multiple distinct periods, such as from binary systems with two spotted stars \citep[e.g.][]{lurie2015} are detected by \citet{mcquillan2014}, but are not included in the following analysis.
-
-The Gaia Data Release 1 (DR1) provides astrometric positions for over $10^9$ sources from the first year of observation with Gaia.
-The Tycho-Gaia Astrometric Solution (TGAS) measures improved proper motions and parallaxes for 2 million nearby, bright sources by extending the astrometric solutions from Tycho and Hipparcos.
-While the TGAS data are not a complete astrometric survey, and have possible systematics in the reported parallaxes \citep{gaia_dr1,stassun2016}, they represent a significant improvement in the astrometry and kinematics available for stars in the \Kepler field thanks to the precision of Gaia.
-
-Using the CDS X-Match service, I cross-matched the available catalogs from these two surveys.
-A default cross-match radius of 5 arcsec was used.
-A total of 33,855 stars were found in cross match between these catalogs, 99.5\% of the sample from \citet{mcquillan2014}.
-A subset of 1,299 objects were recovered in the TGAS sample.
-Due to the brightness limits of the TGAS sample very few K and M dwarfs were recovered in the TGAS sample.
-Future releases of Gaia data will provide full astrometric solutions for nearly all \Kepler stars.
-The rotation periods versus stellar effective temperatures for the \Kepler--Gaia matched stars are shown in Figure \ref{fig:all}.
-
-%tgas = J/ApJS/211/24/table1 \& I/337/tgas
-%gaia = J/ApJS/211/24/table1 \& I/337/gaia
 \begin{figure}[]
@@
 
+%%%%%%%%%%%%%%%%%%%%%%
+\section{The \Kepler--Gaia Data}
+Rotation periods in this study come from \citet{mcquillan2014}, who performed an Auto-Correlation Function analysis of \Kepler stars cooler than 6500 K that had at least $\sim$2 years of observation.
+The periods recovered from this approach generally agree very well with those found via Lomb-Scargle Periodograms \citep[e.g.][]{reinhold2013,aigrain2015}.
+Sources with multiple distinct periods, such as from binary systems with two spotted stars \citep[e.g.][]{lurie2015} are detected by \citet{mcquillan2014}, but are not included in the following analysis.
+
+The Gaia Data Release 1 (DR1) provides astrometric positions for over $10^9$ sources from the first year of observation with Gaia.
+The Tycho-Gaia Astrometric Solution (TGAS) measures improved proper motions and parallaxes for 2 million nearby, bright sources by extending the astrometric solutions from Tycho and Hipparcos.
+While the TGAS data are not a complete astrometric survey, and have possible systematics in the reported parallaxes \citep{gaia_dr1,stassun2016}, they represent a significant improvement in the astrometry and kinematics available for stars in the \Kepler field.
+
+Using the CDS X-Match service, I cross-matched the available catalogs from these two surveys.
+A default cross-match radius of 5 arcseconds was used.
+A total of 33,855 stars were found in the cross match between these catalogs, 99.5\% of the sample from \citet{mcquillan2014}.
+The small number of stars not recovered from \citet{mcquillan2014} may be due to source confusion within the matching radius.
+A subset of 1,299 objects were recovered in the TGAS sample.
+Due to the brightness limits of the TGAS sample very few K and M dwarfs were recovered in the TGAS sample.
+Future releases of Gaia data will provide full astrometric solutions for nearly all \Kepler stars.
+The rotation periods versus stellar effective temperatures for the \Kepler--Gaia matched stars are shown in Figure \ref{fig:all}.
+
 %%%%%
@@
 Two isochrones from \citet{bressan2012} are shown, with ages of 300 Myr and 4 Gyr (solid and dashed lines).
+Main sequence stars are selected as those between the 300 Myr isochrone and the isochrone shifted up by 1 magnitude (red line).
 }
@@
 
-% (gaia[u'parallax_error'] < 0.4) &
-  %            (gaia[u'phot_g_mean_flux_error']/gaia[u'phot_g_mean_flux'] < 0.01))
 Though \citet{mcquillan2014} attempted to only measure periods for dwarf stars, the sample of \Kepler--Gaia matched stars contains both main sequence dwarfs and evolved stars (giants and subgiants).
@@
 Example isochrones from the \citet{bressan2012} grid are shown for two ages.
-A systematic offset between the measured absolute $G$-band and the isochrone main sequence of $\sim$0.5 magnitudes is found (Figure \ref{fig:hist}).
+A systematic offset of $\sim$0.5 magnitudes is found between the measured absolute $G$-band and the isochrone's main sequence.
+This offset is likely due to calibration differences between the nominal and actual $G$-band (A.
+Brown, private communication).
 
 The HR-period diagram in Figure \ref{fig:HR} shows stars with a range of evolutionary states, and could help test post-main sequence angular momentum evolution models \citep[e.g.][]{donascimento2012}.
-Outliers in this diagram (e.g.
-the rapidly rotating star at $T_{eff}=4500$ K, $M_G=4.5$ mag, $P_{rot}$=0.652 d, KIC 07957709) are either due to erroneous cross-matching between the \Kepler and Gaia catalogs, or represent interesting systems such as rare binary star configurations or stars that have undergone mergers or ingested giant planets \citep{massarotti2008,tayar2015}.
-Explaining such outliers is beyond the scope of this work, but these targets are worth further study as they may reveal new physics.
+Outliers in this diagram are either due to erroneous cross-matching in the \Kepler and Gaia catalogs, or represent interesting systems such as rare binary star configurations or stars that have undergone mergers or ingested giant planets \citep{massarotti2008,tayar2015}.
+For example, examining the 2 Micron All Sky Survey \citep[2MASS][]{2mass} image on SIMBAD for
+the rapidly rotating star at $T_{eff}=4500$ K, $M_G=4.5$ mag, $P_{rot}$=0.652 d, in Figure \ref{fig:HR} (KIC 07957709), this source appears highly contaminated with three point sources clustered within $\sim$10 arcseconds, likely leading to an erroneous position in the HR diagram, and possibly incorrect variability measurements with \Kepler.
+Investigating all such outliers in Figure \ref{fig:HR} is beyond the scope of this work, but these targets are worth further study as they may reveal new physics.
 
-The difference between the 300 Myr isochrone main sequence track and the observed $M_G$ is shown in Figure \ref{fig:hist}.
-Given the systematic offset of $\sim$0.5 mag, a fairly wide band of stars ($0\le \Delta M_G \le 1$) was selected as being ``close to the main sequence''.
+Main sequence stars were selected using a simple cut around 300 Myr isochrone.
+Given the systematic offset of $\sim$0.5 mag between the Myr isochrone and the observed $M_G$ values, a fairly wide band of stars ($0\le \Delta M_G \le 1$) was selected as being ``close to the main sequence''.
 The final sample included 440 stars.
@@
 More precise selection will require an improved isochrone track, as well as updated parallaxes from the full Gaia DR2.
-
-%%%%%
-\begin{figure}[]
-\centering
-\includegraphics[width=4in]{fig3}
-\caption{Distribution of differences between the Gaia DR1 $G$ absolute magnitudes and values for main sequence stars from the 300 Myr isochrone shown in Figure \ref{fig:HR}.
-The range of sources selected near the single star main sequence (red lines) may have additional contamination due to dust extinction and binary companions.}
-\label{fig:hist}
-\end{figure}
 
@@
 \centering
-\includegraphics[width=5.3in]{fig4}
+\includegraphics[width=6in]{fig3}
 \caption{ Rotation period versus temperature for TGAS-matched stars near the isochrone main sequence (blue circles). 
@@
 Many different studies have produced competing gyrochrone models, each with unique morphologies at the hot and cool star regimes \citep[e.g.][]{barnes2007,mamajek2008,meibom2011,angus2015}.
-A 600 Myr gyrochrone from \citet{meibom2011} was determined by eye to approximately trace the period minima from 3500 K to 6000 K.
+A 600 Myr gyrochrone from \citet{meibom2011}, converted from $B-V$ color to temperature using the the transformation from \citet{sekiguchi2000}, was determined by eye to approximately trace the period minima from 3500 K to 6000 K.
 As shown in Figure \ref{fig:gyro}, this model (as with most gyrochronology models) turns down in period sharply for stars hotter than $\sim$6000 K.
@@
 \centering
-\includegraphics[width=4in]{fig5}
+\includegraphics[width=4in]{fig4}
 \caption{Residual of log rotation periods about the \citet{meibom2011} 600 Myr gyrochrone, using the log-linear extrapolation between 6000--6500 K shown in Figure \ref{fig:gyro}.
@@
 
+%%%%%
+\begin{figure}[]
+\centering
+\includegraphics[width=4in]{fig5}
+\caption{Total proper motion distributions for stars above (rotating slower, older stars) the gyrochrone model shown in Figure \ref{fig:gyro} (black) and below (rotating faster, younger stars) the model (green).
+Median values for the two distributions are shown (vertical dashed lines), which yield a marginal 2.8$\sigma$ difference.
+}
+\label{fig:pm}
+\end{figure}
+
+\clearpage
+
 The difference (in log period) between the observed rotation and the 600 Myr gyrochrone for the 440 likely main sequence stars is shown in Figure \ref{fig:diff}.
@@
 This is in contrast to the overall results from \citet{mcquillan2013} who found nearly equal numbers of M dwarfs in the fast and slow rotating groups, but is in general agreement with their sample of stars with non-zero proper motions.
-%IDL> print, 262./178.
-%      1.47191 ==> US
-%IDL> print, 345./266.
-%      1.29699 ==> THEM
-
-%%%%%
-\begin{figure}[]
-\centering
-\includegraphics[width=4in]{fig6}
-\caption{Total proper motion distributions for stars above (rotating slower, older stars) the gyrochrone model shown in Figure \ref{fig:gyro} (black) and below (rotating faster, younger stars) the model (green).
-Median values of the two distributions are shown (thick lines), which yield a marginal 2.8$\sigma$ difference.
-}
-\label{fig:pm}
-\end{figure}
 
@@
 As the typical error in the  Gaia DR1 total proper motion is $\sim$1.4 mas/yr for this sample, this is a statistically marginal (2.8$\sigma$) difference.
+Similarly, the Kolmogorov-Smirnov statistic for these two samples is 0.14, which does not rule out the null hypothesis that they are drawn from the same distribution.
+The slower rotating sample (black in Figure \ref{fig:pm}) appears to have a bimodal distribution in proper motion, indicating possibly significant contamination from a younger, lower proper motion population that is consistent with the rapidly rotating stars.
+Further investigation of rotation and kinematics of nearby stars using larger samples from future Gaia data releases is needed.
 
@@
 
-Using a combination of data from \Kepler and Gaia DR1, I have explored the rotation period distribution for 894 nearby main sequence stars.
+Using a combination of data from \Kepler and Gaia DR1, I have explored the rotation period distribution for 440 nearby main sequence stars.
 A bimodal rotation period distribution has been found in stars with temperatures ranging from 5000 K to 6500 K.
@@
 While a definitive explanation for this period bimodality has not been reached, the findings to date seem to favor stellar ages as the cause.
-In this scenario the star formation history for nearby stars would be dominated by two main bouts of star formation, one centered at a few hundred Myr, and one at a few Gyr (slightly younger than the Sun).
+In this scenario the star formation history for nearby stars would be dominated by two epochs of star formation, one short event centered at a few hundred Myr, and one long event centered at a few Gyr (slightly younger than the Sun).
 It is also worth noting that the space volume probed by the TGAS sample investigated here is very similar to that covered by the temperature-selected cool star sample in \citet{mcquillan2013}.
 The median parallax distance for stars in this work is 285 pc, while the median isochrone distance for the K and M dwarfs is $\sim$216 pc.
-This points to the period distribution being a localized age artifact, and explains why previous exhaustive studies of rotation periods in stellar clusters did not find such a bimodality.
+This points to the period distribution being a localized age artifact.
 Determining how localized this age distribution is, and if it can be confirmed for stars across the HR diagram including giants, is a key goal for future Gaia data releases.
@@
 The period bimodality may yet be a manifestation of the ``Vaughan-Preston'' gap observed in chromospheric activity indicators from solar type stars.
-Such a feature has also been discussed for rotating stars by\citet{kado-fong2016}.
+Such a feature has also been discussed for rotating stars by \citet{kado-fong2016}.
 Given that the mass range for the bimodality explored here and in \citet{mcquillan2014} covers stars with solar-type dynamos (those having a tachocline, late F through early M) such a model cannot be fully ruled out at this time.
 Though there have been many rotation studies for cool stars \citep[e.g.][]{irwin2011,newton2016,stelzer2016} too few rotation periods have been measured for stars across the ``fully convective boundary'' ($T_{eff}<3000$ K, spectral type $\sim$M4) to tell if the bimodal period feature continues to cooler temperatures, which would support the age distribution model.
+If the bimodality is due to stars crossing a phase of rapid angular momentum evolution, we would expect to see it in stellar clusters at or near the critical age.
+The lack of this feature in the clusters observed to date could be due to no cluster being close enough to the critical age, which the gyrochrone in Figure \ref{fig:gyro} shows is near 600 Myr.
 Further studies of rotation periods for stars in intermediate age open clusters (e.g.
-the Hyades) may also help solve this mystery \citep[e.g.][]{douglas2014}.
+the Hyades) may help solve this mystery \citep[e.g.][]{douglas2014}.
 
@@
 Looking forward to the astrometric precision of future Gaia data releases, this combination will be effective at separating dwarf stars from subgiants for nearly the entire \Kepler and K2 databases, and enable accurate age maps for field stars.
-% super useful for determining approximate log g values (separating main sequence from sub-giants), and thus more accurate gyrochronology ages \citep{van-saders2013}.
 
@@
 
-This research made use of the cross-match service provided by CDS, Strasbourg.
+This research made use of the cross-match service, and the SIMBAD database, provided by CDS, Strasbourg, France.
 
@@
 \bibitem[{{Gaia Collaboration}(2016)}]{gaia}
-{Gaia Collaboration}. 2016, ArXiv e-prints, arXiv:1609.04153, doi:10.1051/0004-6361/201629272
+{Gaia Collaboration}. 2016, ArXiv e-prints, arXiv:1609.04153
 
@@
 
+\bibitem[{{Sekiguchi} \& {Fukugita}(2000)}]{sekiguchi2000}
+{Sekiguchi}, M., \& {Fukugita}, M. 2000, \aj, 120, 1072
+
+\bibitem[{{Skrutskie} {et~al.}(2006){Skrutskie}, {Cutri}, {Stiening},
+  {Weinberg}, {Schneider}, {Carpenter}, {Beichman}, {Capps}, {Chester},
+  {Elias}, {Huchra}, {Liebert}, {Lonsdale}, {Monet}, {Price}, {Seitzer},
+  {Jarrett}, {Kirkpatrick}, {Gizis}, {Howard}, {Evans}, {Fowler}, {Fullmer},
+  {Hurt}, {Light}, {Kopan}, {Marsh}, {McCallon}, {Tam}, {Van Dyk}, \&
+  {Wheelock}}]{2mass}
+{Skrutskie}, M.~F., {Cutri}, R.~M., {Stiening}, R., {et~al.} 2006, \aj, 131,
+  1163
+
 \bibitem[{{Skumanich}(1972)}]{skumanich1972}

### 2016-11-21 paper/ms.tex
update from referee reply

@@
 \documentclass[manuscript, letterpaper]{aastex6}
+%\documentclass[twocolumn, trackchanges]{aastex6}
 %\documentclass[twocolumn]{aastex6}
@@
 
-%\altaffiltext{1}{Corresponding author: James.Davenport@wwu.edu}
 \altaffiltext{1}{Department of Physics \& Astronomy, Western Washington University, 516 High St., Bellingham, WA 98225, USA}
@@
 
+%tgas = J/ApJS/211/24/table1 \& I/337/tgas
+%gaia = J/ApJS/211/24/table1 \& I/337/gaia
+\begin{figure*}[]
+\centering
+\includegraphics[width=6in]{fig1}
+\caption{
+Rotation period distribution for 33,855 \Kepler stars from \citet{mcquillan2014} with detections in Gaia DR1 (blue dots).
+The period bimodality can be seen most clearly for stars with $T_{eff} < 4000$ K as a dearth of sources with periods of $\sim$25 days, but extends to at least $T_{eff}\sim5500$ K according to \citet{mcquillan2014}.
+The subsample of 1,299 nearby objects found in TGAS are highlighted (red circles), and are mostly hotter stars due to the faint limit of the TGAS sample.
+}
+\label{fig:all}
+\end{figure*}
+
 \citet{mcquillan2013} discovered a bimodal period distribution for M dwarfs in the \Kepler field, which was subsequently confirmed to exist for K dwarfs as well \citep{mcquillan2014}.
@@
 This combined sample allows improved selection of main sequence stars, and reveals previously undetected structure in the rotation period distribution for solar-type stars.
-
-\begin{figure}[]
-\centering
-\includegraphics[width=6in]{fig1}
-\caption{
-Rotation period distribution for 33,855 \Kepler stars from \citet{mcquillan2014} with detections in Gaia DR1 (blue dots).
-The period bimodality can be seen most clearly for stars with $T_{eff} < 4000$ K as a dearth of sources with periods of $\sim$25 days, but extends to at least $T_{eff}\sim5500$ K according to \citet{mcquillan2014}.
-The subsample of 1,299 nearby objects found in TGAS are highlighted (red circles), and are mostly hotter stars due to the faint limit of the TGAS sample.
-}
-\label{fig:all}
-\end{figure}
 
@@
 \centering
-\includegraphics[width=5.3in]{fig2}
+\includegraphics[width=3.5in]{fig2}
 \caption{Hertzsprung--Russell (HR) diagram using temperatures from \citet{mcquillan2014} and Gaia DR1 $G$-band absolute magnitudes for the 894 stars that pass photometric and parallax quality cuts described in the text.
@@
 %%%%%
-\begin{figure}[]
+\begin{figure*}[]
 \centering
@@
 \label{fig:gyro}
-\end{figure}
+\end{figure*}
 
@@
 As shown in Figure \ref{fig:gyro}, this model (as with most gyrochronology models) turns down in period sharply for stars hotter than $\sim$6000 K.
-A log-linear extrapolation of the 600 Myr gyrochrone at 6000 K continues to trace the period bimodality up to 6500 K, and roughly traces a line of constant Rossby number assuming a local convective turnover timescale such as from \citet{barnes_kim2010}.
+A log-linear extrapolation of the 600 Myr gyrochrone at 6000 K continues to trace the period bimodality up to 6500 K, and roughly traces a line of constant Rossby number assuming a local convective turnover timescale such as from \citet{barneskim2010}.
 
@@
 \centering
-\includegraphics[width=4in]{fig4}
-\caption{Residual of log rotation periods about the \citet{meibom2011} 600 Myr gyrochrone, using the log-linear extrapolation between 6000--6500 K shown in Figure \ref{fig:gyro}.
-The bimodal rotation period distribution is clear, with peaks at $\Delta \log P_{rot}$ of -0.19 and 0.21 days.
-A two-Gaussian model was fit to this distribution (blue line).
-Approximately 50\% more stars are present in the slower, nominally older peak (right).}
+\includegraphics[width=3.5in]{fig4}
+\caption{Residual of log rotation periods about the \citet{meibom2011} 600 Myr gyrochrone, using the log-linear extrapolation between 6000--6500 K shown in Figure \ref{fig:gyro} \added{for the 440 main sequence selected stars (thick black line) and the initial sample of 894 stars with good TGAS detections (thin green line).}  The bimodal rotation period distribution \added{for main sequence stars} is clear, with peaks at $\Delta \log P_{rot}$ of -0.19 and 0.21 \replaced{days}{dex} from a two-Gaussian model fit (blue line).
+Approximately 50\% more stars are present in the slower, nominally older peak (right).
+}
 \label{fig:diff}
 \end{figure}
+
+The difference (in log period) between the observed rotation and the 600 Myr gyrochrone \deleted{for the 440 likely main sequence stars} is shown in Figure \ref{fig:diff}.
+Despite combining stars of all temperatures, the bimodality is clearly seen in this log period space \added{for the 440 likely main sequence stars}.
+A two Gaussian model was fit to this data, which found peaks in the two distributions of $-0.19 \pm 0.01$ and $0.21 \pm 0.01$ \replaced{days}{dex}.
+262 stars had periods longer than the gyrochrone model (right peak) and 178 slower than the model (left peak).
+This is in contrast to the overall results from \citet{mcquillan2013} who found nearly equal numbers of M dwarfs in the fast and slow rotating groups, but is in general agreement with their sample of stars with non-zero proper motions.
+\added{The whole sample of 894 stars with good TGAS detections does not show this bimodality in Figure \ref{fig:diff}, demonstrating the importance of culling subgiants from the sample.}
 
@@
 \centering
-\includegraphics[width=4in]{fig5}
-\caption{Total proper motion distributions for stars above (rotating slower, older stars) the gyrochrone model shown in Figure \ref{fig:gyro} (black) and below (rotating faster, younger stars) the model (green).
-Median values for the two distributions are shown (vertical dashed lines), which yield a marginal 2.8$\sigma$ difference.
+\includegraphics[width=3.5in]{fig5a}
+\includegraphics[width=3.5in]{fig5b}
+\caption{Top: Total proper motion distributions for stars above (rotating slower, older stars) the gyrochrone model shown in Figure \ref{fig:gyro} (black), and below (rotating faster, younger stars) the model (green).
+\deleted{Median values for the two distributions are shown (vertical dashed lines), which yield a marginal 2.8$\sigma$ difference.}
+\added{Bottom: Total proper motion as a function of the residual log rotation period shown in Figure \ref{fig:diff}, with the same color scheme for faster and slower rotating stars as the top panel.}
 }
@@
 \end{figure}
-
-\clearpage
-
-The difference (in log period) between the observed rotation and the 600 Myr gyrochrone for the 440 likely main sequence stars is shown in Figure \ref{fig:diff}.
-Despite combining stars of all temperatures, the bimodality is clearly seen in this log period space.
-A two Gaussian model was fit to this data, which found peaks in the two distributions of $-0.19 \pm 0.01$ and $0.21 \pm 0.01$ days.
-262 stars had periods longer than the gyrochrone model (right peak) and 178 slower than the model (left peak).
-This is in contrast to the overall results from \citet{mcquillan2013} who found nearly equal numbers of M dwarfs in the fast and slow rotating groups, but is in general agreement with their sample of stars with non-zero proper motions.
 
@@
 This difference in kinematics versus rotation period is in the same direction observed by \citet{mcquillan2013}.
-As the typical error in the  Gaia DR1 total proper motion is $\sim$1.4 mas/yr for this sample, this is a statistically marginal (2.8$\sigma$) difference.
-Similarly, the Kolmogorov-Smirnov statistic for these two samples is 0.14, which does not rule out the null hypothesis that they are drawn from the same distribution.
+\deleted{As the typical error in the  Gaia DR1 total proper motion is $\sim$1.4 mas/yr for this sample, this is a statistically marginal (2.8$\sigma$) difference.
+Similarly, }The Kolmogorov-Smirnov statistic for these two samples is 0.14, which does not rule out the null hypothesis that they are drawn from the same distribution.
 The slower rotating sample (black in Figure \ref{fig:pm}) appears to have a bimodal distribution in proper motion, indicating possibly significant contamination from a younger, lower proper motion population that is consistent with the rapidly rotating stars.
-Further investigation of rotation and kinematics of nearby stars using larger samples from future Gaia data releases is needed.
+\added{Figure \ref{fig:pm} also shows the proper motion as a function of the residual rotation period defined in Figure \ref{fig:diff}.
+Despite the small sample size, this distribution highlights the complex relationship between ages derived from gyrochronology and galactic kinematics, including bimodal structure for the proper motion of slower rotating stars.} Further investigation of rotation and kinematics of nearby stars using larger samples from future Gaia data releases is needed.
 
@@
 Looking forward to the astrometric precision of future Gaia data releases, this combination will be effective at separating dwarf stars from subgiants for nearly the entire \Kepler and K2 databases, and enable accurate age maps for field stars.
+% super useful for determining approximate log g values (separating main sequence from sub-giants), and thus more accurate gyrochronology ages \citep{van-saders2013}.
 
@@
 Special thanks to Kevin Covey, Jennifer van Saders, Sean Matt, Travis Metcalfe, and Aleks Scholz for their helpful discussions that motivated publishing this work.
+\added{Thanks to the anonymous referee for their helpful comments that improved this manuscript.
+Jay Strader provided a helpful reference for the correct notation for transcendental functions, \citet{matta2011}.}
 JRAD is supported by an NSF Astronomy and Astrophysics Postdoctoral Fellowship under award AST-1501418.
@@
 
+%\bibliography{/Users/davenpj3/Dropbox/references.bib}
 \begin{thebibliography}{}
@@
 
-\bibitem[{{Barnes} \& {Kim}(2010)}]{barnes_kim2010}
+\bibitem[{{Barnes} \& {Kim}(2010)}]{barneskim2010}
 {Barnes}, S.~A., \& {Kim}, Y.-C. 2010, \apj, 721, 675
@@
 
+\bibitem[{{Matta} {et~al.}(2011){Matta}, {Massa}, {Gubskaya}, \&
+  {Knoll}}]{matta2011}
+{Matta}, C.~F., {Massa}, L., {Gubskaya}, A.~V., \& {Knoll}, E. 2011, Journal of
+  Chemical Education, 88, 67
+
 \bibitem[{{McQuillan} {et~al.}(2013){McQuillan}, {Aigrain}, \&
